---
title: "Correlation and Regression—Moving Together"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Foundational Concepts

In everyday life, social situations, you have lots of examples of cause
and effect. If a student studies more, that student may possibly do
better on the next test. The income of a person is related to the amount
of education that they have. But you could certainly ask the question of
whether the increase in the income is a result of the knowledge or
skills they gained from education.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)
library(broom)
library(RColorBrewer)


```

```{r}
  def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats
```

```{r}

cmStats$μ(1:10)
Xs = c(0, 5, 20, 3, 15, 8)
Ys = c(2, 10, 10, 6, 25, 4)

sort(Xs)[(6 + 1) / 2]
median(sort(Xs))


cmStats$P(1, 6)
```

Statistical graphs are useful in helping us visualize data. Through
graphs, we understand data properties, such as the mean, median, and
standard deviation; find patterns in data, such as clustering and
correlation; suggest an underlying model that could have generated the
data; verify our assumptions; check and fortify our analysis; summarize;
and communicate results. This lecture will define and identify basic
summaries of data, both numerical and graphical, and use R for
calculating descriptive statistics, making graphs, and even writing
functions that work on multiple datasets.

```{r}
 ggplot(satgpa, aes(x = sat_sum, fy_gpa)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Verbal SAT percentile",
    y = "First year (college) grade point average"
  )
```

Describing relationships and cause and effect among variables is a basic
strategy for understanding the world. A statistical challenge is to
describe and measure how two variables, such as incoming SAT scores and
college GPAs, are related. Higher SAT scores correspond with better GPAs
in college; however, the SAT scores do not cause higher grades. In this
lecture, we will introduce the ideas of scatter plots, which give a
visual sense of the relationships between two variables; correlation,
which gives a quantitative measure of the strength of the linear
relationship between two varying quantities; and linear regression,
which produces a straight-line approximation for a set of paired
variables.

```{r}
cor(satgpa$sat_sum, satgpa$fy_gpa)
```

We can quantify the extent to which two variables are related by giving
a number defining the correlation that summarizes the relationship. If
one quantity increases and the other quantity tends to increase, the
correlation is called positive, and the opposite is negative. Unlike the
scatter plot of SAT scores and GPAs, which demonstrates a positive
correlation, the scatter plot of two negatively correlated quantities
goes down as it goes to the right.

```{r}
cor(satgpa$sat_v, satgpa$sat_m)
```

```{r}
 ggplot(satgpa, aes(x = sat_sum, hs_gpa)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Verbal SAT percentile",
    y = "First year (college) grade point average"
  )
```

```{r}
cor(satgpa$hs_gpa, satgpa$fy_gpa)
```

The correlation is denoted r (or [the Greek letter rho] ). It is
computed as follows: For each member of the population (e.g., student),
take the distance of the rst variable (SAT score) from the rst
variable’s mean in units of standard deviation (that is, take the
z-score of the SAT score) multiplied by the distance of the second
variable (GPA) from its mean, again measured in units of standard
deviation (that is, multiply by the z-score of the GPA).

```{r}
cmStats$rho_sym(X, Y)
```

We often summarize a scatter plot by a straight line. If the data set
has a high correlation, the straight line will lie near most of the
points. The straight line is a kind of summary of the data. The line is
called a linear regression line. It is a line that approximates the
data.

```{r}
c(sd(satgpa$sat_sum), mean(satgpa$sat_sum))

```

```{r}
c(sd(satgpa$fy_gpa), mean(satgpa$fy_gpa))
```

Similar mechanisms are used when there are more than two variables.
Several explanatory variables can often make better predictions than any
single variable alone. When two or more variables are used to predict
another variable, we use multiple regression to summarize the situation.

```{r}
cor(satgpa)
```

Statistical correlation indicates an association but does not prove that
there is a causal relationship between the variables. **One of the
misuses of statistical information is to mistakenly infer cause and
effect from correlation.**

# Concept and Applications

## VALIDATING STATISTICAL ASSUMPTIONS

Statistical graphs are useful in helping us visualize data. Through
graphs, we understand data properties, such as the mean, median, and
standard deviation; find patterns in data, such as clustering and
correlation; suggest an underlying model that could have generated the
data; verify our assumptions; check and fortify our analysis; summarize;
and communicate results. This lecture will define and identify basic
summaries of data, both numerical and graphical, and use R for
calculating descriptive statistics, making graphs, and even writing
functions that work on multiple datasets.

### BAR PLOTS

```{r}
attach(iris)
```

The “iris” dataset is widely used throughout statistical science for
illustrating various problems in statistical graphics, multivariate
statistics, and machine learning. It’s a small but nontrivial dataset.
The data values are real ( as opposed to simulated ) and are of high
quality ( collected with minimal error ). The data were used by the
celebrated British statistician Ronald Fisher in 1936.

```{r}
ggplot( iris, aes(x = Species, y = Petal.Length)) +
  geom_col()


ggplot( iris, aes(x = Species, y = Sepal.Length)) +
  geom_col()

ggplot(iris, aes(x = Sepal.Length, colour =Species)) +
  geom_his() +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )
```

### BOX PLOTS

The summary function is a quick and easy way to assess the statistical
properties of each attribute. These values are displayed graphically in
a box plot.

```{r}
summary(iris[,1: 2])
```

```{r}
summary(iris[,3: 4])

```

Box plots are used to compactly show many pieces of information about a
variable’s distribution. They are great for visualizing the spread of
the data. Box plots show 5 statistically important numbers: the minimum,
the 25th percentile, the median, the 75th percentile, and the maximum.

```{r}
iris_long <- iris |> tidyr::pivot_longer(cols = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), names_to= "Measurement", values_to= "Value")

ggplot(iris_long, aes(x = Measurement, y = Value)) +
  geom_boxplot()

ggplot(iris_long, aes(x = Measurement, y = Value, fill = Species)) +
  geom_boxplot()
```

### SCATTERPLOTS

Scatterplots are helpful for visualizing data and simple data
inspection. Let’s try the following code

Let’s generate corresponding scatterplots for Petal.Width, Sepal.
Length, and Sepal.Width.

```{r}

ggplot(iris, aes(x = Species, y = Petal.Length)) +
  geom_point(alpha = 0.6, position = position_jitter(width = 0.2))

ggplot(iris_long, aes(x = Measurement, y = Value, fill = Species)) +
  geom_point(alpha = 0.6, position = position_jitter(width = 0.2))
```

It continues in this fashion for Petal.Length and Petal.Width. Bring
your attention to the bottom-left corner, the last plot in column 1.
This plot has Sepal.Length on the 𝑥-axis and Petal.Width on the 𝑦-axis.

```{r}

# 1. Calculate the correlation matrix
cor_matrix <- cor(iris[, 1:4])

# 2. Reshape the data for ggplot2
# First, convert the matrix to a data frame and add row names as a column
cor_df <- as.data.frame(cor_matrix)
cor_df <- tibble::rownames_to_column(cor_df, var = "Var1")

# Then, use pivot_longer() to reshape it
cor_long <- tidyr::pivot_longer(
  cor_df,
  cols = -Var1,
  names_to = "Var2",
  values_to = "Correlation"
)

# 3. Create the heatmap plot
ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "red", high = "green", mid = "white",
    midpoint = 0, limit = c(-1, 1),
    name = "Pearson\nCorrelation"
  ) +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 4) +
  labs(
    title = "Correlation Matrix of Iris Measurements",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r}
names(iris[, 1:4])
```

### HISTOGRAMS

A histogram is a plot that breaks the data into bins ( or breaks ) and
shows the frequency distribution of those bins.

Let’s create some histograms of our iris data. The number of bins in the
histogram is variable.

```{r}
ggplot(iris, aes(x = Petal.Width)) +
  geom_histogram(bins = 13) +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )

ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(bins = 25) +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )

```

### DENSITY PLOTS

Density plots can be viewed as smoothed versions of a histogram. We can
estimate the density using R’s density function.

```{r}
ggplot(iris, aes(x = Petal.Width)) +
  geom_density() +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )

ggplot(iris, aes(x = Petal.Length)) +
  geom_density() +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )

ggplot(iris, aes(x = Sepal.Width)) +
  geom_density() +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )

ggplot(iris, aes(x = Sepal.Length)) +
  geom_density() +
  labs(
    x = "Sepal lenght",
    y = "Frequecy"
  )
```

Let’s also look at the density function of Petal.Length for each of the
3 classes of irises.

```{r}
ggplot(iris, aes(x = Petal.Length, fill = Species)) +
  geom_density()
```

### CONTOUR PLOTS

Density estimation is available for higher-dimensional data using
contour plots. A contour plot is a graph that explores the potential
relationship among 3 variables.

Contour plots display the 3-dimensional relationship in 2 dimensions,
with 𝑥 and 𝑦 variables plotted on the 𝑥 and 𝑦 scales and the z variable
represented by contours.

```{r}

# Create the contour plot using geom_density_2d()
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d(aes(color = Species)) +
  labs(
    title = "Contour Plot of Iris Data Density",
    subtitle = "Traces show the density of points for each species",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)"
  )

```

The plot may also be viewed as a heat map, with brighter colors denoting
more values in those regions.

```{r}
# Create separate grid heatmaps for each species
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_density_2d_filled(aes(fill = after_stat(level))) +
  labs(
    title = "Grid Heatmap by Species",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)"
  )
```
