---
title: "Law—You’re the Jury"
output: html_notebook
---

# Foundational Concepts

[C]ertainly, in the law, there are many places where statistical data and
inferences are used to help in making legal decisions. In this lecture,
we’re going to look at two cases where you will be the jury.
```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)

```

```{r}
def_sym("X", "Y", "R")

cmStats <- modules::use("CmStats.R")$cmStats

```

We’ll look at two main examples of courtroom drama where we
imagine ourselves on the jury. We’ll present the facts, and it will
be up to us to determine what conclusions are appropriate to
draw. We begin with a hit-and-run accident.

This example illustrates a surprising statistical
anomaly known as Simpson’s Paradox. 

We will present the evidence; you determine the verdict.

• Mr. Jones witnessed a hit-and-run accident involving a cab.

• Mr. Jones stated that he thought the cab was blue.

• The jury has come to the conclusion that a guilty verdict hangs on
whether or not Mr. Jones’s testimony implies that the probability that
the cab was blue is over 50%.

• We did some experiments to determine how accurate Mr. Jones’s vision
was. He was able to correctly identify the blue cabs as blue 80% of
the time and the green cabs as green 80% of the time. The prosecutor
summarized the situation by saying that Mr. Jones is 80% sure the cab
was blue; therefore, the jury should convict.

Some additional information was presented:

• There are exactly 100 cabs in the city.
• 90 are green cabs and 10 are blue.

```{r}
cabs_n = 100
blue_cabs = 10
green_cabs = 90
```

Let’s do the following thought experiment: Suppose instead of having one
accident, all 100 cabs in the city simulated this accident. Each one did it
once. How many times would Mr. Jones have testified that the cab was blue?
Following the logic, we see that he would testify “blue” 26 times, but only 8
of those times would the cab actually be blue.

```{r}
cmStats$P(blue_cabs, cabs_n) 
cmStats$P(green_cabs, cabs_n)

```

Another case that demonstrates the same statistical issue about testing for rare
events arises when considering random drug testing in a company. Suppose
the company has 280,000 employees, of whom 500 employees actually use
the illegal drugs that are the target of this policy. Suppose the drug test will
correctly read positive for 95% of people who actually use those drugs; thus,
475 employees who use drugs would receive a positive test result. 

```{r}
employees_n = 280000
under_substances = 500
tested_positive = 475
```

Now suppose that the test gives a false positive
1% of the time; of the 279,500 employees
who do not use drugs, 1% will get a false-
positive result. That is, the drug test will
read positive for 2795 employees who do
not, in fact, use drugs.

```{r}
employees_n - under_substances
cmStats$P(under_substances, employees_n)
```

```{r}
employees_n - tested_positive
cmStats$P(tested_positive, employees_n)
```
```{r}
employees_n - under_substances - tested_positive
cmStats$P(employees_n - under_substances - tested_positive, employees_n)
```

The total number of employees whose drug
tests are positive will be 475 + 2795 = 3270
employees.

```{r}
tested_positive + (employees_n - tested_positive)
```

Therefore, if an employee gets
a positive test result, his or her chance of actually using drugs is 475/3270,
or less than 15%. Thus, there is a great danger of inappropriate rings or
accusations based on positive drug testing.

```{r}
cmStats$P(tested_positive, tested_positive + (employees_n - tested_positive))

```

We now turn to gender discrimination. To determine discrimination by
gender, race, religion, or age, it is natural to look at data about treatment
of specific groups. Let’s consider a case of admissions to a program at a university.
For the purposes of this example, let’s assume that 1000 men
applied and 1000 women applied to the program and that all 2000 applicants
had exactly the same qualifications. Here are the facts:

```{r}
mans <- 1000
womens <- 1000

applicants_n = mans + womens
```

There was a 70% acceptance rate for the men and only a 40% acceptance
rate for the women. It appears that the women were clearly discriminated
against, and of course, we are outraged. The chi-square test can identify the
acceptance rate you would expect from random chance alone.

```{r}
excellent_app_n = (mans - 800) + (womens - 200)

mediocre_app_n = (mans - 200) + (womens - 800)

```

But let’s look at the case more deeply. Suppose the total program to which
the 2000 applied actually had two subprograms. One subprogram was an
Excellent Program to which 200 men applied and 800 women applied. The
other subprogram was a Mediocre Program to which 800 men applied and 200
women applied. Of the 200 men who applied to the Excellent Program, 20%
were accepted. Of the 800 women who applied to the Excellent Program, 25%
were accepted.
```{r}
cmStats$P((mans - 800), excellent_app_n) + .20
```
```{r}
cmStats$P((womens - 200), excellent_app_n) + .25

```
In the Mediocre Program, 800 men applied and 82.5% were accepted,
while 200 women applied to the Mediocre Program and 100%
were accepted. In each subprogram, a higher percentage of women were
accepted. This situation is an anomaly, because overall, it appeared that
the women were being discriminated against; however, looking at the
subprograms we see a different picture.

```{r}
cmStats$P((mans - 200), mediocre_app_n) + 82.5

```

```{r}
(cmStats$P((womens - 800), mediocre_app_n) + 1.0)

```
This scenario is an illustration of a phenomenon known as Simpson’s
Paradox, that is, a situation where both subprograms indicate that the
women are being treated better, yet overall, the men appear to be treated
better. Given that the women actually had a higher acceptance rate in each
subprogram, let’s think about how we would decide whether the differences
in acceptance rates are serious enough to be viewed as clear discrimination
or whether the differences could be reasonably accounted for as the result of
simple random luck.


Let’s look at a table that records the data:

```{r}
excellent_app_df <- tibble(
  applicants = c("women", "men", "total"), 
  accept = c(40, 200, 240),
  reject = c(160, 600, 760),
  total = c(200, 800, 1000),
  accept_rates = c(.20, .25, .24)
    )
excellent_app_df

mediocre_app_df <- tibble(
  applicants = c("women", "men", "total"), 
  accept = c(660, 200, 860),
  reject = c(140, 0, 140),
  total = c(800, 200, 1000),
  accept_rates = c(.82, 1.0, .86)
    )
mediocre_app_df
```
How rare an event would it be to have one acceptance rate as much as 5%
less than the other? To determine how surprised we should be at getting a 5%
difference in acceptance rates, we can look at a normal probability curve that
tells us the probability of different deviations.
The gender discrimination case illustrated a statistical anomaly known as
Simpson’s Paradox. Simpson’s Paradox is an example of a possible effect
of a lurking variable. In that case, the lurking variable was the existence of
the subprograms.

In this lecture, we saw how a witness’s testimony was not as compelling
as it at rst appeared. We saw that universal or random drug tests present
problems because of false positives. In the gender-discrimination case, we
saw an example of Simpson’s Paradox; we also saw how the principles
of hypothesis testing helped us to interpret evidence as indicative of
discriminatory practices or not.
# Concept and Applications

## LINEAR REGRESSION MODELS AND ASSUMPTIONS

n regression, we look at the association between 2 or
more quantitative variables. We’re going to begin to
look for whether one variable causes or predicts changes
in another variable. The response variable, which is the
dependent variable, might measure an outcome of a study
or experiment. The explanatory variable, which is the
independent or predictor variable, explains or is related
to changes in the response variable. We now have pairs
of observations: $( 𝑥1, 𝑦1 ), ( 𝑥2, 𝑦2 ), …, ( 𝑥𝑛, 𝑦𝑛 )$. In linear
regression, we’re looking for the association between
2 variables to be centered on a line. In essence, we’re
looking for the effect that one variable has on another.

### LINEAR REGRESSION

Imagine that we are wheat farmers in Kansas. For the past 10 years, at
the end of the season, we’ve recorded the total amount of rainfall and
the average height of our wheat. We’d like to know if rainfall has any
effect on our wheat yield.

We have 10 data points, each recording the amount of rainfall ( measured
in inches ) and wheat growth ( in inches ). Here’s our data in R:

```{r}
rainfall = c(3.07,3.55,3.90,4.38,
4.79,5.30,5.42,5.99,6.45,6.77)
wheat = c(78,82,85,91,92,96,97,104,111,119)
summary(cbind(rainfall, wheat))
```
Notice that our scatterplot shows that there might be a linear
relationship between rainfall and wheat yield.

```{r}
plot(rainfall, wheat)
```
We can summarize the relationship between the 2 variables by
fitting a regression line, which is a straight line that describes how a
response variable 𝑦 changes as an explanatory variable 𝑥 changes and
minimizes the squared difference ( also called residuals ) between each
observation and the fitted line.

Our linear regression model, with unknown parameters, looks like this:
$𝑦 = β0 + β1𝑥 + 𝜖$, where $β0$ is the 𝑦-intercept, $β1$ is the slope, and $𝜖$ is the
error term.

R will conveniently fit a linear regression for us. We use the function
“lm,” which stands for linear regression. After “lm” comes our response
variable, “wheat,” followed by a tilde, which tells R that we want to
model wheat on the explanatory variables that follow it. In this case,
that’s rainfall.

```{r}
reg <-  lm(formula = wheat ~ rainfall)
reg
```
With the summary statistics for the residuals, we can quickly see if
our residuals have any outliers and check to see that the median is
close to 0.

“Coefficients” is where we find our estimates: 0 = 45.0408 ( the
𝑦-intercept term ) and 1 = 10.1691 ( our slope coefficient for wheat ).

The following is a plot of our wheat data with the estimated regression
line.

```{r}
qqnorm(reg$residuals)
qqline(reg$residuals, lwd=2, col="red")
```
### PREDICTIONS

We can use a regression line to predict the value of the response
variable 𝑦 for a specific value of the explanatory variable 𝑥. This value is
called a predicted value or fitted value.

Be careful about extrapolations. We can’t conclude that the regression
line is an accurate model outside our sample data. In other words, just
because our data supports a linear relationship between 𝑦 and 𝑥 doesn’t
mean that that relationship holds for values outside the range of 𝑥.
Predictions of 𝑦 for values of 𝑥 that are far away from the actual data
range often aren’t accurate.

Let’s use our regression equation to actually
predict the wheat yield.

```{r}
ŷ = reg$coefficients[1] + reg$coefficients[2] * X
ŷ
```

With 5 inches of rain:

```{r}
ŷ = reg$coefficients[1] + reg$coefficients[2] * 5
ŷ
```
With 100 inches of rain:

```{r}
ŷ = reg$coefficients[1] + reg$coefficients[2] * 100
ŷ
```

### RESIDUALS

We need a way of understanding how well our regression line fits the
data. The residuals help us do just that. A residual is the difference
between an observed value $( Yi )$ and the value predicted by the
regression line $( Ŷi ): ei = 𝑦i − ŷi$.


Some people get confused at this point, because it seems like the
residuals, $e_i$, are the same as our errors, $𝜖_i$.


𝜖 measures the deviation of the observed value $( 𝑦 ) $from the
unobservable, true regression line.
ۧ ۧ The residuals, on the other hand, measure the difference between the
observed value $( 𝑦i )$ and the estimated regression line $( ŷi )$.

A good plot, with some great residuals, looks like total randomness. There’s no pattern. And approximately the same number of points
will be above and below the $e = 0$ line. A bad plot has some obvious pattern, such as a funnel shape or parabola. There might be more points above $0$ than below ( or vice versa ).

The “no problem” graph is an example of a good residual plot. The data values are nicely centered at $0$, and the points are well balanced above and below the line at $𝑦 = 0$.

The following is a residual plot of the
wheat data. There appears to be a
pattern, but because there are so few
data points, it’s best if we look at a
quantile-quantile ( Q-Q ) plot—which
confirms that our data, even though
it’s small, satisfies the normality
assumption.

```{r}
rainfall = c(3.07,3.55,3.90,4.38,4.79,
5.30,5.42,5.99,6.45,6.77)
wheat = c(78,82,85,91,92,96,97,104,111,119)
wheat.lm = lm(wheat~rainfall)
plot(wheat.lm$fitted.values,wheat.lm$residuals,
main = "Residuals vs. Fitted Values",
xlab = "Fitted Wheat Values", ylab = "Residuals")
abline(h=0, col="red")
```
```{r}
qqnorm(residuals(lm(wheat~rainfall)))
qqline(residuals(lm(wheat~rainfall)), lwd=2, col="Blue")

```
### ASSESSING MODEL FIT

Now that we’ve fit a linear model to our data, we need to figure out how good it is. How well does it fit our data? Is the model strong enough that
we could get good predictions from it? The following terminology will
help us asses model fit.

The regression sum of squares ( SSR ) measures the variation in ŷi.

```{r}
SSR <-  sum_((cmStats$y_hat - X)^2, cmStats$i, 1, cmStats$n, doit = FALSE)
SSR
```
The error sum of squares ( SSE ) measures the residual variation in 𝑦
due to error. This is the sum of our residuals squared.

```{r}
SSE <-  sum_((Y - cmStats$y_hat)^2, cmStats$i, 1, cmStats$n, doit = FALSE)
SSE
sum_(cmStats$e^2, cmStats$i, 1, cmStats$n, doit = FALSE)

```
The total sum of squares ( SST ) measures the total variation in 𝑦.

```{r}
SST <-  sum_((cmStats$y - Y)^2, cmStats$i, 1, cmStats$n, doit = FALSE)

SST
```
Note: $SST = SSR + SSE$.
The coefficient of determination, R2, is the fraction of the variation in 𝑦
that can be explained by the linear regression of 𝑦 on 𝑥.

```{r}
SSR + SSE
```

```{r}
R_sqr <- SSR/SST
R_sqr

1 * (SSE/SST)

```
An R2 of 0 means that the 𝑦 cannot be predicted linearly from 𝑥. An
R2 of 1 means that the 𝑦 can be predicted without error from 𝑥. An R2
between 0 and 1 indicates the extent to which 𝑦 is predictable by 𝑥. ( An
R2 of 0.10 means that 10% of the variance in 𝑦 is predictable from 𝑥; an
R2 of 0.80 means that 80% is predictable. )


We can’t forget to check our model
assumptions.

To check for normality, we look at a
normal Q-Q plot of the residuals.

To check for constant variability, we
check the residual plot.

To check for independence, we
examine the way in which subjects/
units were selected in the study or
how the data was collected.

To check for linearity, we check a scatterplot or a residual plot.

It’s always important to check that the assumptions of the regression
model have been met; otherwise, our results won’t be valid.

