---
title: "Hypothesis Testing—Innocent Until"
output: html_notebook
---

# Foundational Concepts

Hypothesis testing is one of the workhorses of statistical inference. The
goal of this lecture is crystal clear: to illustrate and explain the logic
of hypothesis testing. Basically, that’s the core of the idea of statistical
inference in general.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas);
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)

```

```{r}
def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats;
```
Hypothesis testing is one of the most common statistical techniques
used in education, psychology, social sciences, natural sciences,
medicine, law, and every area to which statistics is applied.

We’ll rst describe how hypothesis testing works in an example. Our null
hypothesis is that if we spin a penny, it has a .5 probability of landing heads
up. At the end of the whole hypothesis testing process, the crux of the decision
of whether to reject the null hypothesis or not depends on quantifying how
rare different possible outcomes of the experiment are.

Thus, we need to investigate the possible outcomes of the coin experiment
and how likely or unlikely each possible outcome is. We will spin the penny
100 times.


```{r}
coins <- sample(c("T", "H"), size=100, replace = TRUE)

cmStats$P(table(coins), 100)
```

```{r}
coins_df = tibble(flips = coins)
```

```{r}
ggplot( coins_df, aes(y = length(coins_df), x = flips)) +
  geom_col()
```
Under our hypothesis, all possible outcomes are equally likely
because each spin’s landing is independent of that of the other spins. We
can make a graph in which the x-axis (the horizontal axis) tells how many
heads are in the outcome and the y-axis tells how many of all the possible
outcomes have exactly that many heads.

Next, we do the experiment and analyze the results. We spin the coin 100
times.

```{r}
coins_df = tibble(flips = sample(c("T", "H"), size=100, replace = TRUE))

```

```{r}
ggplot( coins_df, aes(y = length(coins_df), x = flips)) +
  geom_col()
```
```{r}
count_table = table(coins_df)
count_table
```

```{r}
cmStats$P(count_table, 95)
```


The probability of getting
that value or a value further from the mean is called the p value. It is the
probability that under the assumption of the null
hypothesis, an outcome as rare as (or rarer than)
the actual outcome would happen. The smaller
the p value, the more extreme the outcome.
Our intuition that the result of the experiment
is strong evidence against the null hypothesis
is quanti ed and con rmed by the fact that the
p value is very small.

```{r}

t.test(count_table)# not realy correct for the text below.
```
Under the assumption of the hypothesis, the
computations show that the probability of getting
the experimental results is very low; thus, the
experimental results are strong evidence for rejecting the null hypothesis.
The phrase reject the null hypothesis is the way of saying that the hypothesis
about the world is unlikely to be true.


Another example of hypothesis testing occurs if we are trying to determine
whether or not a particular medication works by giving some people the
medication and some a placebo and seeing whether there is a signi cant
difference in the number who recover. For our experiment, we’ll take 100
people, all with athlete’s foot, and have them use the medication for a week.
After the week, we count how many people are free from athlete’s foot.

Assume that we know that 40% of the time, a person’s athlete’s foot will
disappear in a week without any medication. Thus, our null hypothesis is that
each person has a 40% probability of cure with or without the medication
(i.e., the medication has no effect). We’ll take as the alternative hypothesis
that the probability of cure for people using the medication is different from
40%. We also decide how extreme a result of the experiment will have to be
for us to reject the null hypothesis.

```{r}
p_value = 0.05
prob_null = 0.40
```
Our required signi cance level is 0.05, meaning that if the results of the
experiment are so extreme that under the assumption of the null hypothesis,
only 5 out of 100 experiments would yield a result that extreme or more
extreme, we will reject the null hypothesis.

```{r}
p_value ~ cmStats$P(5, 100)
```
```{r}
cmStats$P(5, 100)
```
In other words, if the p value is
less than the signi cance level, we will reject the null hypothesis. We call
such a result statistically signi cant. In the eld of statistics, signi cant does
not mean “important.” Rather, it simply means that the signi cant item or
event signi es something.

```{r}
free_foot = 51
```

Suppose that 51 of the 100 people who used the medication are free from
athlete’s foot. In this case, the calculation gives the p value of

```{r}
(free_foot/100) - (prob_null / 2) + 0.01
```
the calculation gives the p value of 0.032, quite
small. We conclude that it is very unlikely that the medication had no effect.
Because the p value of 0.032 is less than the signi cance level of 0.05 that
we speci ed before we did the experiment, we reject the null hypothesis.


As another example, consider the question of whether the typical American
male adult consumes an average of 2400 calories each day. We ask 25
American adult males, and they average 2500.

```{r}
people = 25
avg_cal = 2400
reported_cal = 2500
std_cal = 270
```

Unlike the previous examples, knowing just the sample mean is not
enough information. We also need to know something about how spread
out the population’s eating habits are, that is, the standard deviation of the
population. If there is a huge variation from person to person in number of
calories consumed, then the difference between the sample mean of 2500
and the null hypothesis of 2400 is not as signi cant. The sample responses
themselves give some information about the standard deviation of the
population. Let’s suppose that the sample standard deviation (i.e., the spread
in the 25 responses) is 270 calories.

```{r}
mans_cal_2400 = rnorm(1:people, mean = avg_cal)
mans_cal_2500 = rnorm(1:people, mean = reported_cal, sd = std_cal)

```

```{r}
t.test(mans_cal_2500, mu = reported_cal) # this conclusion needs to be review!
```
The general idea is exactly the same as in the previous examples. We do
a calculation, based on the t distribution, that determines how extreme our
sample is, assuming the null hypothesis is true. It turns out that a sample
of mean 2500 calories and standard deviation 270 calories is not extreme,
assuming the null hypothesis of 2400 calories and requiring a 0.05 level of
significance. The p value is 0.085, greater than 0.05. We do not reject the
null hypothesis. That is, the data would not warrant a conclusion that the
average number of calories consumed by an American adult male is different
from 2400.

State a hypothesis about the way the world is. The hypothesis is called
the null hypothesis and is often the opposite of what we actually think
might be true.

• State an alternative hypothesis. In the case where the null hypothesis
is that the probability of something is a certain value, three standard
possible alternative hypotheses exist:

• The probability is different from the value stated in the
null hypothesis.

• The probability is greater than the value stated in the
null hypothesis.

• The probability is less than the value stated in the null hypothesis.

• Declare what level of signi cance we require to reject the
null hypothesis.

• Do the experiment, gather the results, and compute the probability that,
under the assumption of the null hypothesis, the results would be as
extreme as (or more extreme than) the result we actually obtained. This
gives a p value.

• If the result is rare (p value smaller than our required level of
signi cance), then we assert that the results are statistically signi cant,
and we reject the null hypothesis:

• Rejecting the null hypothesis means concluding that it is false.

• Rejecting the null hypothesis also means concluding that the
alternative hypothesis is true.

• If the result is not rare (p value larger than our required level of
signi cance), then we do not reject the null hypothesis.
# Concept and Applications

## INTERVAL ESTIMATES AND CONFIDENCE INTERVALS

When we calculate a point estimate, our chances
of hitting the target population parameter are
not very likely. We’ll often get close but will
seldom hit the mark. In this lecture, you will learn about
confidence intervals, with which we can increase our
chances of capturing the true population parameter.

### CONFIDENCE INTERVALS

A confidence interval draws inferences about a population by estimating
the value of an unknown parameter using an interval. We’re looking
for an interval that covers the true population parameter with some
amount of certainty or confidence.

But the problem is that a different random sample ( with different
observed values ) would yield different estimates for μ. Which of those
estimates ( the sample mean from our first sample or the sample mean
from our second sample ) would be closest to the true value? We’d really
have no way of knowing.

To create a confidence interval, we start with the point estimate ±
estimation error. If our interval is narrow, that means that our estimate
is precise.

Let X1, X2, … X𝑛 be a random sample from a normal distribution with
μ and σ. We know that regardless of sample size 𝑛, X is normally
distributed with expected value μ and standard deviation . We get
this from the central limit theorem.

```{r}
cmStats$Z_score_sym(X)
```
```{r}
Z = 0.95 ~ P(-1.96 < Z, Z < 1.96)
```

We’ve just found a random interval.

```{r}
confiden_interval <-  list(cmStats$X - 1.96 * cmStats$sigma/cmStats$sqrt_sym(cmStats$n), cmStats$X + 1.96 * cmStats$sigma/cmStats$sqrt_sym(cmStats$n))

confiden_interval # centered at X.
```
### CONFIDENCE LEVELS

Confidence intervals allow us to estimate population parameters using a range of values—a range of values that are more likely to capture the true population parameter.

To do that, we need to set the confidence level, which tells us how likely it
is that the population parameter is actually contained in the confidence
interval. It’s a measure of the degree of the reliability of the interval.

In statistics, we typically use a confidence level of 95%, but we could
specify any confidence level as ( 1 − α )100%. Note that when α = 0.05,
we get the 95% confidence level. This means that 95 out of 100 samples
( 95% ) from the same population will result in a confidence interval
that contains the true population parameter.
```{r}
(1 - 0.05) * 100 / 100
```

### 95% CONFIDENCE INTERVALS

After observing X1 = 𝑥1, X2 = 𝑥2, …, X𝑛 = 𝑥𝑛 from a normal distribution
with known σ, we compute the sample mean X . The 95% confidence
interval for the population mean μ is

```{r}
confiden_interval
```
We can also say with 95% confidence,
```{r}
list(confiden_interval[1], cmStats$mu, confiden_interval[2]) # mu is between the interval.
```
Suppose that a normal population has unknown μ and σ and X = 80.0, what is the 95% confidence interval for μ?

```{r}
subs(cmStats$X - 1.96 * cmStats$sigma/cmStats$sqrt_sym(cmStats$n), list(X = 80, n = 31, σ = 2.0))
```
```{r}
-0.0632258064516129 * sqrt(31) * 2.0 + 80
    
```

```{r}
subs(cmStats$X + 1.96 * cmStats$sigma/cmStats$sqrt_sym(cmStats$n), list(X = 80, n = 31, σ = 2.0))
```
```{r}
0.0632258064516129 * sqrt(31) * 2.0 + 80
```
Given a 95% confidence interval, is it correct to say that μ falls in the
confidence interval with probability 0.95?

When we substitute X with the observed value, x , there’s no randomness
left. For a particular sample, either μ is in the interval or it isn’t.

A precise way to interpret confidence interval is that with 95%
confidence, the interval calculated includes μ. It’s the confidence
interval itself that is random ( i.e., it changes for each sample ); μ is just
a fixed number ( the population mean ). 

It is nonrandom, but we usually
don’t know what it is. That’s why we’re taking a sample.
ۧ ۧ The 95% means that we used a method that captures the true mean
95% of the time.

Let’s define zα/2 as our critical value. This is the value such that
$P( Z >zα/2 ) = P( Z < −zα/2 ) = α/2, where Z ~ N( 0, 1 )$.

### OTHER CONFIDENCE LEVELS

We’re not limited to a 95% confidence interval. We can also specify
other alpha ( α ) levels.

 When α equals 0.1, that gives us 90% confidence. We have to divide
0.1 by 2 to get the area that’s left in the tails.
 
 Likewise, when α equals 0.05, we get 95% confidence and 2.5 % left
in each of the tails.
 
 When α equals 0.01, we get 99% confidence and 0.5% area in each of
the tails.

```{r}
list(cmStats$X - cmStats$Z_alpha/2 * cmStats$sigma/cmStats$sqrt_sym(cmStats$n), cmStats$X + cmStats$Z_alpha/2 * cmStats$sigma/cmStats$sqrt_sym(cmStats$n))
```
Several of the most commonly used zα/2’s are as follows.
```{r}
tibble(
  level = c("alpha/2" , "Z_alpha/2"), 
  "90%" = c(0.05, qnorm(0.05)), 
  "95%" = c(0.025, qnorm(0.025)), 
  "99%" = c(0.005, qnorm(0.005)) 
)

```
### POINT ESTIMATES AND CONFIDENCE

The Organic Milk Association produces organic gallon milk for more
than half of the United States. The company claims that, due to
manufacturing upgrades, they are confident that the average amount
of milk in each gallon container they ship has a volume of 128 ounces.

We can use R to simulate a population of amounts of milk in 10,000 milk
jugs. We assume some distribution function and randomly create
10,000 data points in a variable called “milk.”

```{r}
set.seed(343)
milk = 129 - rexp(100000,0.95)
hist(milk, main="Histogram
of Milk Population",
col="red")
```
The histogram sample is skewed
on the left ( meaning that the
data is piled up on the right ).

```{r}
true_mean = mean(milk)
true_sd = sd(milk)
true_mean
```
```{r}
true_sd
```
We can take a sample from our population to see how close we are to
the actual true mean.
```{r}
set.seed(343)
sample_milk <-  sample(milk, size = 50, rep = T)
sample_mean <- mean(sample_milk)
sample_mean
```

```{r}
sample_mean - true_mean
```
Our sample mean is only slightly larger than the population mean, by
0.04 ounces.

We can calculate the 95% confidence interval for the sample mean in R.

```{r}
n = 50
sample_milk <-  sample(milk, size = n, rep = T)
sample_mean <- mean(sample_milk)
sample_mean - 1.96 * sd(sample_milk) / sqrt(n)
```

```{r}
sample_mean + 1.96 * sd(sample_milk) / sqrt(n)

```

This is a box plot of our data, with population on the left and sample on
the right.

```{r}
boxplot(milk, sample_milk)
```
Our point estimate based on a sample of just 50 milk jugs only slightly
overestimates the true population. This illustrates an important point:
We can get a fairly accurate estimate of a large population by sampling a
relatively small subset of individuals.

But what might the central limit theorem tell us about this population?
Remember, the central limit theorem says that no matter what the
shape of the population distribution, the shape of the distribution of
sample means is approximately normal.

Suppose that we take 1000 samples of size 𝑛 = 50 and look at the sample
distribution.

```{r}
milk_mean <- mapply(\(v) mean(sample(milk, n, rep=T)), 1:1000)

hist(milk_mean)

```

The distribution of sample means looks more like a normal distribution
than the distribution of the population.

How normal is the distribution of 1000 sample means? One way to
quantify this is to make the normal quantile-quantile ( Q-Q ) plot
comparing the quantiles of our distribution against the quantiles of a
truly normal distribution.

```{r}
qqnorm(milk_mean)
qqline(milk_mean)
```
We want to compute some confidence intervals from our samples.
In this ( artificial ) case, we know the population parameters: μ ( the
population mean ) and σ ( the population standard deviation ). We can
use R to calculate the confidence intervals.

Now let’s return to our samples of size 𝑛 = 50 and take 100 such samples.

```{r}

sampe_mean <- mapply(\(v) mean(sample(milk, n, rep=T)), 1:100)
hist(samp_mean)

```
```{r}
m = 20; n = 50; mu = mean(milk);
sigma = sd(milk);
SE = sigma/sqrt(n)
alpha = 0.10 ; zcrit = qnorm(1-alpha/2);
matplot(rbind(samp_mean[1: m] - zcrit*SE, samp_mean[1: m]
+ zcrit*SE),
rbind(1: m,1: m),
type="l", lty=1,lwd=2,
col='darkgray',
xlab = "Ounces",
ylab="Confidence Intervals",
main="90% Conf. Int.");
abline(v=mu)
```

The 80% confidence interval
has shorter lines that the 90%
confidence interval and is less
likely to cover the true
population mean.

Note that we use qnorm( 0.975 )
to get the desired z critical
value instead of qnorm( 0.95 )
because the distribution has
2 tails.
