---
title: "Quack Medicine, Good Hospitals, and Dieting"
output: html_notebook
---

# Foundational Concepts

Every day, we make decisions about our personal health; it’s one of the
most important things that we think about. Our state of health is often
measured by data. The kind of data that we are interested in are things
like our weight, our blood pressure, the levels of our cholesterol, and
other various chemicals in our blood stream. These data present to us a
picture of our well-being.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)
library(sp)
library(spdep)
#library(rgeos) # Might issue a deprecation warning, but is needed for old workflows
#library(geoR)
library(gstat)

# We still use the modern sf package to reliably access the data
library(sf)
library(spData)
#library(maptools) #

```

```{r}
def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats

```
We make decisions every day about our personal health. If our
cholesterol is above some number, our doctor is likely to suggest
a medication to supplement healthful eating and exercise. Such
a recommendation is commonly based on statistical results of studies that
show a correlation between high cholesterol and increased risk of heart
attack. Aspects to consider in applying results of such a study to ourselves
include how like us the people in the study are and the difference between
correlation and causation. Another statistical concept, regression to the mean,
explains why quack medicine may appear to work often.

### Condition "personalize medicine?"

What we would really like is a study involving people who are as much
like us as possible, because their experience with, say, a medication is more
apt to be similar to ours. We’d like to condition the overall study data on
several variables, looking at the subset of
the study data that matches us with respect
to those variables. In general, the concept
of conditioning the data on some criteria
means that we look at only the data with a
certain feature. We prefer to respond to the
statistical results conducted on people most
similar to us so that we would have a sense
that the studies pertained most specifically
to us.

Tailoring the statistics to our individual personal health situation could
conceivably provide a greater increase in personal health than would
improvements in treatments themselves. Of course, doctors are aware that
different patients react differently, and to some extent, they try to tailor their
treatments to the individual. Using their experience to modify the results of
the studies can be good, but it can also be problematic, because an individual
doctor sees many fewer patients than there are in some studies.

Another statistical issue arises when we need to decide which of two
hospitals to go to for heart surgery. Suppose we have the following chart
summarizing how successful each hospital is with each of three subcategories
of patients: those entering in fair condition, in serious condition, and in
critical condition.
```{r}
tibble(
  patient = c("fair", "serius", "critical", "total"), 
  Hospital_A = c(700, 200, 100, 1000), 
  Hospital_B = c(100, 200, 700, 1000), 
  Survivors_from_A = c(.600/.85, .100/.50, .10/.10, .710/.71), 
  Survivors_from_B = c(.90/.90, .150/.75, .300/.43, .540/.54), 
  )
```
Looking at the data broken down in this way, we see that B has a higher
success rate in all three categories of dif culties. When averaged all together,
however, the impression is that hospital A is superior, with a 71% survival
rate. But hospital B is superior in each of the three categories. Notice that
mathematically, this is the same kind of example as one we saw in an earlier
lecture on gender discrimination. This is Simpson’s Paradox.

### Regression to the mean

Another statistical question arises when trying to distinguish between a
quack medicine and a real medicine. The reason that quack medicines
sometimes seem to work is that usually people recover from a minor disease
even without medicine. Quack medicines appear to work because of the
phenomenon called regression to the mean: An ill person usually expects to
return to his or her mean health situation.

A common issue in the realm of personal health management concerns dieting
and weight. Many quantities, including weight, have an innate variability. If
you weigh yourself daily and chart the results, you’ll notice that the readings
can differ by a pound or two, even if you didn’t “gain” or “lose” weight. If
you weigh yourself to the nearest half a pound, then make a histogram over
the readings, you get a somewhat normal-shaped picture distributed around
a central value. A weight chart is a time series, a value for different times. In
the case of a person losing weight, the weight chart shows a downward trend,
but it does not always go down uniformly each day. We can summarize the
chart by drawing a straight line, the least squares regression line. The weight
loss data illustrate how we use a mathematical model (the straight line) that
captures the spirit and look of the data to summarize the data.

Several statistical issues are related to everyday health issues. We discussed
the notion of conditioning on subpopulations that match a person of interest,
which can give different results than analysis over the whole population.
We illustrated Simpson’s Paradox by the example of choosing a hospital.
We illustrated how regression to the mean can be the explanation for quack
medicine seeming to work and for punishment to work but praise not to
work. We saw how to summarize a time series by a straight line.

# Concept and Applications

## SPATIAL STATISTICS

We tend to think of data as being fixed in time and
space, collected in the moment to be analyzed in
the moment. But what happens when our data
is on the move—when values of our variables change over
space? We can use a branch of statistics called spatial
statistics, which extends traditional statistics to support
the analysis of geographic data. It gives us techniques to
analyze spatial patterns and measure spatial relationships.

### SPATIAL STATISTICS

Spatial statistics is designed specifically for use with spatial data—
with geographic data. These methods actually use space ( area, length,
direction, orientation, or some notion of how the features in a dataset
interact with each other ), and space is right in the statistics. That’s
what makes spatial statistics ( also known as geostatistics ) different
from traditional statistical methods.

.1 Descriptive spatial statistics is similar to traditional descriptive
statistics. For example, if we have lots of points on the map, we might
want to know where the center of those points is located. That’s
equivalent to computing the mean. We might also want to know
how spread out those points are around the center. That’s similar to
computing the standard deviation.

.2 Spatial pattern analysis tells us whether there is any ( nonrandom )
structure to the data. For example: Are features clustered? Are they
uniform? Or are they dispersed in a regular ( and nonrandom ) way?

.3 Spatial analysis is used to measure and evaluate spatial relationships.
Imagine that we are looking at a hot-spot map for 911 calls. We
might be curious about why we are seeing so many calls, or hot
spots, in certain locations. Spatial regression analysis could be used
( the output looks just like regular regression, except you test for
autocorrelation with a test called Moran’s I ). Spatial regression could
identify the factors promoting the spatial pattern we’re observing,
which might help us explain why 911 rates are so high.

polygons. For example, customers might be points, roads could be lines,
and zoning districts could be polygons. In R, any of that information can
be handled in a vector format or a raster format.

Here is data for 100 counties in North Carolina that includes the counts
of live births and deaths due to sudden infant death syndrome ( SIDS )
for 2 periods: July 1974–June 1978 and July 1979–June 1984.

```{r}
spData::nc.sids
```

This map has the latitude and longitude locations for the North Carolina
counties along with other data from the late 1970s and early 1980s.

Who’s your neighbor? How do you decide if somebody is close to you?
And how much influence should they have on your predicted value?

```{r}


if (require(sp, quietly = TRUE)) {
# convert to SpatialPolygonsDataFrame
spdf <- as_Spatial(nc)
# identical to
spdf <- as(nc, "Spatial")
# convert to SpatialPolygons
as(st_geometry(nc), "Spatial")
# back to sf
as(spdf, "sf")
}

# ---------------------------------------
spdf <- as_Spatial(nc)
 

# 3. Define the Breaks
# Using the SIDS Count variable from 1974
 breaks <- quantile(spdf$SID74, probs = seq(0, 1, length.out = 6))

# 4. Plot the Map using spplot
spplot(
  spdf,
  zcol = "SID74",                                  # The variable to map
  col.regions = gray.colors(5),                   # Color scheme (adjust as needed, e.g., "Reds")
  at = breaks,                                      # Use the defined breaks
  main = "North Carolina SIDS Counts, 1974 (sp plot)",
  colorkey = TRUE
)

```
```{r}
par(mar = c(0, 0, 1, 0))
plot(spdf)
invisible(text(coordinates(spdf), labels=as.character(spdf$NAME), cex=0.4))
```
```{r}
library(spdep)

# 1. Define Queen Neighbors (Most common method)
# queen = TRUE means sharing any border or vertex is enough.
nb_queen <- poly2nb(spdf, queen = TRUE)

# 2. Define Rook Neighbors (More restrictive)
# queen = FALSE (the default) means only sharing a full border segment.
nb_rook <- poly2nb(spdf, queen = FALSE)

# Check the resulting structure (a list where each element lists the neighbor indices)
print(nb_queen) 

# Summarize the neighbor relationships
summary(nb_queen)
```

```{r}
plot(nb_rook, 
     coordinates(spdf), # Provide the coordinates (centroids) for drawing the links
     col = "blue",       # Color the links red
     lwd = 1.5          # Set line width
)
invisible(text(coordinates(spdf), labels=as.character(spdf$NAME), cex=0.4))

```
```{r}
plot(nb_queen, 
     coordinates(spdf), # Provide the coordinates (centroids) for drawing the links
     col = "red",       # Color the links red
     lwd = 1.5          # Set line width
)
invisible(text(coordinates(spdf), labels=as.character(spdf$NAME), cex=0.4))
```
```{r}
library(spdep)

# 1. Get the coordinates (centroids) of the polygons
county_centers <- coordinates(spdf)

# 2. Find the K-Nearest Neighbors (K=3 in this case)
# The knearneigh function finds the indices of the K closest points for each point.
knn_nb_list <- knearneigh(county_centers, k = 3)

# 3. Convert the KNN list to a standard 'nb' neighbor list
# This is required for further analysis (like Moran's I) or plotting.
nb_knn <- knn2nb(knn_nb_list)

# Check the structure: it lists the indices of the 3 closest neighbors for each county.
print(nb_knn)

# Summarize the neighbor relationships
summary(nb_knn)
```
```{r}
# Plot the county boundaries
plot(spdf, border = "gray", main = "North Carolina 3-Nearest Neighbors")

# Overlay the KNN links (lines connect the 3 closest counties)
plot(nb_knn, 
     county_centers, 
     col = "blue",      # Use a different color to distinguish from Queen
     lwd = 1.5         
)
```


