---
title: "Election Problems and Engine Failure"
output: html_notebook
---

# Foundational Concepts

[I]n this lecture, we’re going to start out with a case study, where we
look at the preferences of the people and ask the question of “Who is
the people’s choice?” This example actually is a real election—it was
real data from an important election—so it will be interesting for you
to see it and make your own decision.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)

```

```{r}
def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats

```

This lecture begins as a continuation of the previous lecture by looking
at some famous real elections in which we might wonder whether the
will of the people prevailed. The challenge of choosing an election
winner can be thought of as taking voters’ rank orderings of the candidates
and returning a societal rank ordering.

Here is a chart that summarizes the voting preferences of the population,
giving for each candidate the percentage of the population that ranked the
candidate rst, second, and so on.

```{r}
tibble(
  candidate = c("A", "B", "C", "D", "Total_%"),
  "%_of_1st" = c(40, 13, 18, 29, 100),
  "%_of_2st" = c(14, 46, 18, 22, 100),
  "%_of_1+2" = c(54, 59, 36, 51, 200),
  "%_of_3st" = c(16, 33, 3, 48, 100),
  "%_of_1+2+3" = c(70, 92, 39, 99, 300),
  )
```
What follows is a summary of how well each candidate would have done
under each voting scheme.

• In plurality voting, A would win.
• In the vote-for-two scheme, B would win.
• Using a run-off, D would win.
• Using the Borda Count method, D would win.

What candidate do you think best represents the will of the people? Many
people feel that Candidate D is the best choice. Before telling you who
actually won this election, let me tell you that this election was a presidential
election; thus, there is one more column to be added to the table, namely, the
Electoral College votes. In the Electoral College, Candidate A won handily
with 180 electoral votes. Interestingly Candidate C, who was not a contender
in any of the other schemes presented, actually received the second highest
number of electoral votes.

Perhaps now is the time to tell you what election this was: It was the
presidential election of 1860. Candidate A is Abraham Lincoln, B is Bell, C
is Breckinridge, and D is Douglas. The percentages of people who rated the
candidates in second and third places are estimates obtained from historians
72
of the Civil War. It is intriguing to think about the consequences of the
statistical issue of how to summarize the data of the voters’ opinions.

```{r}
tibble(
  candidate = c("A", "B", "C", "D", "Winner"),
  plurality = c(.40, .13, .18, .29, "A"),
  Vote_for_two = c(.27, .30, .18, .25, "B"),
  run_off = c(.46, 0., 0., .54, "D"),
  borda_count = c(164, 165, 92, 179, "D"),
  electoral_college = c(180, 39, 72, 12, "A"),
  )
```
We have seen lots of bad news about election problems. Now, if possible, it
gets worse. Next, we will discuss the Condorcet Paradox. Let’s look at an
example of an election among three candidates, A, B, and C. The following
chart summarizes the views of the 30 voters about these candidates.

```{r}
tibble(rank = c("1_st", "2_st", "3_st"), voters_f_10 = c("A", "B", "C"),
       voters_s_10 = c("B", "C", "A"), voters_t_10 = c("C", "A", "B"))
```

For every candidate, two-thirds of the people have a speci c alternative
candidate whom they prefer. We can produce even worse cases—with 10
candidates, for example, when no matter who is declared the winner, there is
a specific alternative candidate who is preferred by 90% of the voters.


Many people feel that if there is a speci c candidate who would beat every
other candidate in a head-to-head contest, then that candidate, the Condorcet
winner, should be declared the victor. Marie Jean Antoine Nicolas Caritat,
73
marquis de Condorcet, wanted to point out a weakness of the Borda Count
method. He did so by producing the following famous voters’ profile:

```{r}
tibble(rank = c("1_st", "2_st", "3_st"), "30" = c("A", "B", "C"),
       "10" = c("B", "C", "A"), "10_s" = c("C", "A", "B"), "1" = c("A", "C", "B"),
       "29" = c("B", "A", "C"), "1_s" = c("C", "B", "A")
       )
```
B wins with the Borda Count. But A is the Condorcet winner. Thus, the
Borda Count method does not always select the Condorcet winner. However,
in a sort of voting theory double-reverse, recently, Donald Saari has pointed
out an interesting further analysis of this old example. If we rst erased
collections of voters whose votes canceled one another, then B should win.
Perhaps the Borda Count method chose the right winner.

```{r}
tibble("20" = c("A", "B", "C"), "0" = c("B", "C", "A"),
       "0_s" = c("C", "A", "B"), "0_t" = c("A", "C", "B"), "28" = c("B", "A", "C"),
       "0_fo" = c("C", "B", "A"))
```
This example shows again the subtleties of summarizing data meaningfully.

All the examples in the last two lectures suggest that summaries of complex
situations require contextual arguments to decide among them. Statistical
and logical analysis can help a great deal in choosing which arguments to
find most persuasive and which systems to use in which settings. Voting
theory is an intriguing topic for further study.

# Concept and Applications

## MULTIPLE LINEAR REGRESSION

When we first started examining the relationship
between 2 variables, we used t-tests, which
allowed us to determine if there was a
statistically significant relationship between 2 variables.
We then moved to simple linear regression, which allowed
us to fit a line to a predictor versus response. While these
are useful in the case where we only have 2 variables, it’s
more often the case to work with data that has multiple
predictors. This lecture is about multiple linear regression.

### MULTIPLE LINEAR REGRESSION

Multiple regression is a statistical tool that allows us to examine how
multiple predictor variables can be used to model a response variable.

We’ve generalized the simple linear regression methodology to now
describe the relationship between a response variable 𝑦 and a set of
predictorsX1, X2, …, X𝑛 in terms of a linear function.

The variables X1, X2, …, X𝑝 are the 𝑝 explanatory variables ( because they
explain what’s happening ) in the response variable Y.

Y = β0 + β1X1 + β2X2 + … + βpXp + 𝜖, where β0, β1, …, βp are the coefficients
( parameters ), X1, X2,


As in the case of simple linear regression, we want to find the equation
of the line that best fits the data. In this case, it means finding ( 0, 1, …,
p ) such that the fitted values, ŷi, are as close as possible to the observed
values, 𝑦i.
The following data was collected on a population of women 21 and older
of Pima Indian heritage who were living near Phoenix, Arizona. The
women were tested for diabetes according to World Health Organization
criteria. This is a very rich dataset with the following columns:

- npreg: number of pregnancies;
- glu: plasma glucose concentration in an oral glucose tolerance test;
- bp: diastolic blood pressure ( mm Hg );
- skin: triceps skinfold thickness ( mm );
- bmi: body mass index ( weight in kg/( height in m )2 );
- ped: diabetes pedigree function ( information on relatives with diabetes );
- age: age in years; and
- type: “Yes” or “No” for diabetic according to WHO criteria.

```{r}
head(Pima.tr)
```
```{r}
pima = Pima.tr
summary(pima)
```

All of the variables are quantitative except for type, which is categorical
( “Yes” for diseased or “No” for non-diseased ).

```{r}
hist(pima$bp)
```

```{r}
plot(density(pima$bp), main= "Blood Pressure")
```

In the following pairs plot of 4 of the data values, there doesn’t seem to be
a strong relationship in the data values. In fact, skin is a possible outlier.

```{r}
pairs(pima[1:4])
```

```{r}
pairs(pima[5:8])
```
In the following pairs plot of the other 4 quantitative values, there is a
slight relationship between bmi and ped.

```{r}
round(cor(pima[1:7]), 2)
```
We can do a correlation analysis of our data to tell us the relationship
between our variables. Notice that most of the correlations are at or
below 0.34. The highest is between bmi and skin, at 0.66. Then, age
and number of pregnancies is correlated at 0.6. That tells us that these
variables would go nicely into a linear regression fit. We’re not going to
see a strong correlation among them.

Let’s fit a multiple linear regression model to bmi from the rest of the explanatory variables.

```{r}
lm1 <-  lm(bmi ~ npreg + glu + bp + skin + ped + age + type, data = pima)
summary(lm1)
```

We could go through and think hard about which variables to include,
or we could automate the process. Stepwise regression lets us drop
insignificant variables one by one. This is particularly useful if you have
many potential explanatory variables.

Not all of our 𝑥i variables are significant predictors. 
We can drop the most insignificant paremeter out one at a time using the “drop1”
command in R.

```{r}
lm2 <- lm(bmi ~ npreg + glu + bp + skin + ped + age + type, data = pima)
drop1(lm2, test="F") 
```
We can also model glucose as a function of our other predictor variables.

```{r}
lm3 <-  lm(glu ~ bmi + npreg + bp + skin + ped + age + type, data = pima)
summary(lm3)
```
Plot the actual values against the predicted values. Informally, the
model does not appear to be doing a good job.

```{r}
plot(lm3$fitted.values, pima$glu)
predict(lm3)
```
We can check for linearity by examining the residual plot.

```{r}
plot(lm3$fitted.values, lm3$residuals)
```
Check for normality and constant variance. Do the conditions appear to be approximately satisfied for this model?

```{r}
hist(lm3$residuals)
qqnorm(lm3$residuals)
qqline(lm3$residuals)
```
### DATA WITH HIGHER CORRELATION

The Pima dataset is nice because it has several quantitative predictors
and the predictors aren’t highly correlated. But what happens when we
have predictors that are correlated?

Let’s return to the Motor Trend cars data and fit a multiple regression model
by including more explanatory variables in a linear regression model.

Let’s load our “mtcars” dataset.

```{r}
mpg_model <-  lm(mpg ~ wt + hp, data=mtcars)
summary(mpg_model)
```
Let’s add the “horsepower” variable to our original model on the mpg
dataset. Now our proposed regression model is 𝑦 = β0 + β1weight +
β2horsepower + 𝜖, where β0 is our 𝑦-intercept, β1 is our regression
coefficient for weight, and β2 is our regression coefficient for
horsepower.

The coefficient for each of our explanatory variables is the predicted
change in 𝑦 for 1 unit change in 𝑥, given that the other explanatory
variables are held fixed.

For example, as weight increases by 1 unit, the miles per gallon will
decrease by 3.877. Intuitively, this makes sense: The heavier a car gets,
the lower miles per gallon it would tend to get.

The 𝑝-value for each coefficient tells us whether it’s a significant
predictor of 𝑦 given the other explanatory variables in the model.
Notice that the 𝑝-values for both weight and horsepower are significant.

If explanatory variables are associated with each other, coefficients and
𝑝-values will change depending on what else is included in the model.

```{r}
cor(mtcars$wt, mtcars$hp)
```

We see that “hp” and “wt” are correlated with a correlation coefficient
of 0.658. It’s possible that these predictors are collinear, meaning that
they’re either highly correlated or they contribute the same information
to the model.

Because we have several predictor variables in our “mtcars” data, let’s
just throw them all in the model. Perhaps this will improve our model
even more.

```{r}
mpg_model = lm(mpg ~., data = mtcars)
summary(mpg_model)
```
None of our variables are significant. The all have 𝑝-values greater than
0.05. In fact, a model using all the variables doesn’t perform as well as
our weight-and-horsepower–based model. Our adjusted R2 actually
decreased—from 0.8148 to 0.8066.

Adding more explanatory variables will only make R2 increase. The
more predictors we have in the model, the more they consume our
regression sum of squares. Because R2 is the ratio of the regression
sum of squares to the total sum of squares, it will always increase with
additional predictor variables.

One problem in our data is that our variables are correlated. We can see
this in the pairs plots.

```{r}
pairs(mtcars[, c(1, 3: 4)])
```

```{r}
pairs(mtcars[, c( 5: 7)])
```
```{r}
pairs(mtcars[,c(1,3: 7)])
```
We have highly correlated variables. This leads to model misspecification.

```{r}
round(cor(mtcars[, c(1:3, 7)]), 2)
```

In fact, it’s possible to overfit a model by including too many explanatory
variables. **We have to use the principle of parsimony—the simplest,
most efficient model is the best—because the fewer the coefficients
have to estimate, the better they will be estimated.**

We can find the best model by pruning. We “step” through the predictor
variables and remove the ones that are not significant.

### CHOOSING THE BEST MODEL

How do we choose a best model? There are several methods we could
use based on which model features are most important to us.

*Choose the model with the highest adjusted R2.* This assumes that we
choose to evaluate the success of our model in terms of the percentage of
the variability in the response explained by the explanatory variables.

*The 𝑝-value for an explanatory variable can be taken as a rough measure
for how helpful that explanatory variable is to the model.* Insignificant
variables may be pruned from the model as long as adjusted R2 doesn’t
decrease. You can also look at relationships between explanatory
variables; if 2 are strongly associated, perhaps both are not necessary.

Let’s do a stepwise regression on our linear model fit of miles per gallon
with all of our data. This will automatically spit out the best model.

```{r}
mpg_model2 <-  step(lm(mpg ~., data = mtcars))
```
The output from our stepwise regression model is as follows. This is the
best model.

```{r}
summary(mpg_model2)
```
