---
title: "Title"
output: html_notebook
---

# Foundational Concepts

The foundation of statistical inference is comparing the results that we
nd from actual data that we collect with what we should expect from
random processes. But our intuition about what we are supposed to
expect from a random process is often far from accurate.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)
library(broom)
library(RColorBrewer)

```

```{r}
def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats

```

When every possible outcome of an experiment is equally likely, it is easy
to calculate the probability of an outcome. If a coin is equally likely to land
heads as tails when ipped, we say the probability of heads is 50% (or 0.5).
For a fair die, the probability is 1/6 of rolling a 4, for example. For cases
where possible outcomes are not equally likely, the probability of a possible
outcome means the fraction of the time that the outcome will occur.

```{r}
coins <- sample(c("T", "H"), size=100, replace = TRUE)

cmStats$P(table(coins), 100)
```
```{r}
dice <- sample(c(1:6), size=100, replace = TRUE)

cmStats$P(table(dice), 100)

```

Randomness becomes more subtle the more we think about it. When ipping
a coin, in what sense is the outcome a random event? Once we let the coin
go, physics determines what the outcome will be. Interestingly, in quantum
physics, the basic model of particles is inherently probabilistic.


We humans have bad intuition about probabilities. The following is an
example that is surprising to most people. Suppose there are 50 random
people in a room. What is the probability that two of them will have the
same birthday? The surprising answer is that there is a 97% chance that two
of them will have the same birthday.
```{r}
start_day <- 366 - 50 + 1

prod(mapply(\(v) cmStats$P(v, 366), start_day:366))
```

The product of all the fractions is about 0.03. Thus, the probability that no
two people have the same birthday is only about 3%. Hence, the chance that
two people do have the same birthday is about 97%.


When two fair dice are thrown and summed, not all possible sums are equally
likely. Suppose you have a red die and a black die, and each is a fair die.
There are six ways of getting the sum 7: red 1, black 6; red 2, black 5; and
so on. We can draw a histogram showing how many ways there are to get
each of the numbers 2 through 12. The histogram has a peak over the value
7. When three fair dice are thrown and summed, the histogram of possible
sums is more peaked than the histogram for two dice.
```{r}
purrr::map(1:6, \(d) sum(c(d, 7 - d)))
```
```{r}
(7 - 1) / prod(c(6, 6))
```
```{r}
cmStats$P(1, 6)
```
```{r}
dice_red <- sample(c(1:6), size=100, replace = TRUE)
dice_black <- sample(c(1:6), size=100, replace = TRUE)

sum(outer(dice_red, dice_black, FUN = "+") == 7)

```
```{r}
counts <- table(purrr::map2_dbl(dice_red, dice_black, \(x, y) x + y))
counts
```

```{r}
hist(counts)
```
Saying that the histogram is more
peaked is saying that the standard
deviation is smaller. In fact, the
standard deviation for the case of
100 dice is 1/10 of the standard
deviation for the case of a single die. If we used 1 million dice, we would get
an extremely peaked histogram: The standard deviation would be only 1/1000
of the single-die case. The standard deviation decreases proportionately as
the square root of the number of dice increases. The fact that the histogram
becomes more peaked as the number of thrown dice increases is really an
illustration of the central limit theorem.

```{r}
dice_red <- sample(c(1:6), size=1000, replace = TRUE)
dice_black <- sample(c(1:6), size=1000, replace = TRUE)

sum(outer(dice_red, dice_black, FUN = "+") == 7)

```
```{r}
counts <- table(purrr::map2_dbl(dice_red, dice_black, \(x, y) x + y))
counts
```

```{r}
hist(counts)
```

The central limit theorem implies that for a
fairly large sample size n, the distribution of the sample means is close to
the normal distribution centered at the same mean as the population mean
and with a standard deviation approximately equal to the standard deviation
of the original distribution divided by the square root of n. The central limit
theorem is one of the workhorses of statistical inference.

**These sample means have a distribution that is
approximated by a normal distribution.**

In the next lectures, we will see how our ability to understand probabilities
lies at the heart of the logic of statistical inference.

# Concept and Applications

## SAMPLE SIZE AND SAMPLING DISTRIBUTIONS

Statisticians are often called on to do consulting,
whether for individuals, companies, or nonprofits.
This lecture focuses on a consulting example
involving a large metropolitan city that is considering
building a new hospital to meet the needs of residents.
They want to survey the population to better understand
the typical emergency room ( ER ) demand so that
they can plan an appropriate number of beds.

### SAMPLE MEANS

One thousand residents are sampled and their number of visits to the
ۧ
ER are recorded. The number of times a person visited the ER ranges
from 0 to 59 times in a year. The counts for each visit are given in
the vector “counts.” The vector “visits” repeats each “time” for each
corresponding “count.”

```{r}
times <- (0: 59)

counts <- c(49, 51, 50, 85, 47, 61, 29, 29, 32, 21, 36,
38, 30, 27, 24, 34, 38, 37, 24, 32, 26, 23, 27, 22, 19,
10, 10, 12, 13, 8, 12, 5, 6, 6, 1, 11, 2, 4, 2, 1, 3, 0,
0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1)

visits <- rep.int(times, counts)

```


The following histogram displays information about the distribution
of visits for our population of 1000 residents. Notice the shape of the
distribution.

```{r}
hist(visits)
```
If we look at different samples of residents and their mean number of
visits, how would the estimate change? What if our sample size was
100? 10,000? Let’s investigate how the distribution of a collection of
sample means changes as the sample size increases.

```{r}
mean.age <- function(n){
  trials = 1000
  my.samples <- matrix(sample(visits, size = n * trials, replace = TRUE), trials)
  
  means <- apply(my.samples, 1, mean)
  means
}
```

The function mean.age( 𝑛 ) performs 1000 times the experiment of
drawing a sample of visits of size 𝑛 at random with replacement from
our dataset. This is called bootstrapping a sample from data.

```{r}
par(mfrow=c(1, 2))
means_variances = c(1, 2, 10, 20, 100, 200)

plots_mean_er =function(m) {
  hist(mean.age(m), main = sprintf("Mean of  %s, Visit", m), xlab="Number of Visits")
  hist(visits, main="ER Visits Data ", xlab=" Number of Visits")
}

mapply(plots_mean_er, means_variances)
```

### VARIANCE OF X N

We see from the histograms that the distribution of the sample mean
X 𝑛 becomes narrower as 𝑛 increases. In other words, the variance
decreases as the sample size gets larger.
ۧ ۧ Let’s see if there’s a functional relationship between the variance
distribution of the sample mean X 𝑛 and the sample size 𝑛. We can
calculate the variances of mean.age( 1 ), mean.age( 2 ), mean.age( 10 ),
mean.age(20), mean.age( 100 ), and mean.age( 200 ) with a function
called “vars.”

```{r}
variances <- data.frame(n=means_variances, variance = mapply(\(m) var(mean.age(m)), means_variances))
variances
```

Let’s look at the relationship of variance with sample size 𝑛.
```{r}
plot(variances$n, variances$variance)
```
In addition to becoming narrower, the shape of X 𝑛’s distribution
changes from being skewed to becoming symmetric and bell-shaped,
like the normal distribution, as 𝑛 becomes larger.

### QUANTILE-QUANTILE PLOTS

We can test how close the distribution of X 𝑛 is to the normal distribution
by examining quantile-quantile ( Q-Q ) plots. In a Q-Q plot, quantiles of
the sample are plotted against quantiles of a proposed distribution,
also known as theoretical quantiles.

```{r}
par(mfrow=c(2,3))
mapply(\(m) qqnorm(mean.age(m)), means_variances)

```
If the points of a Q-Q plot fall on a straight line, this indicates that the
sample data are consistent with the proposed distribution ( here, the
normal distribution ). As 𝑛 increases, what changes in the Q-Q plots?

### SHAPIRO-WILK TEST

```{r}
par(mfrow=c(2,3))
mapply(\(m) shapiro.test(mean.age(m)), means_variances)

```

### SAMPLING DISTRIBUTIONS

A statistic is a value computed from data. In our example, our statistic is
the average number of ER visits.

Is a statistic a random variable? Recall that a random variable is a
numeric value associated with the outcome of an experiment. It has a
distribution, a mean ( expected value ), and a variance.

Because data constitute a random realization of an experiment, each
time we run the experiment, we will get a different value for our statistic.
So, in fact, it has a distribution, a mean ( expected value ), and a variance.

**In other words, as our sample size increases, any distribution—normal
or not—will have a mean that tends to behave normally.**

### THE CENTRAL LIMIT THEOREM

In the field of statistics, it’s rarely possible to collect all of the data from
an entire population. Instead, we can gather a subset of data from a
population and then use statistics for that sample to draw conclusions
about the population.

The 2 most common characteristics that we use to define a population
are its mean and standard deviation. And when data comes from a
normal distribution, we immediately have an idea of the data’s shape.

The mean tells us the center of that distribution. The standard deviation
tells us the spread. 

**The central limit theorem tells us that, no matter
what the population distribution looks like, the distribution of the
sample means will approach a normal distribution.**