---
title: "Samples—The Few, The Chosen"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Foundational Concepts

In this lecture, we’re going to come to grips with the question of what
only part of the data really means. So, this lecture is about the
concept of a sample.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)

```

```{r}
def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats

```

With this lecture, we begin the investigation of how to infer features
of the whole population from information about just some of the members
of the population. A common scenario in which this plays out occurs when
a poll is taken to estimate what proportion of voters favor which
candidate.

Recall that the structure of our course is to view statistical analysis
as having two basic parts: (1) how to describe, summarize, and organize
a collection of data if we know all the data and (2) how to infer
information about the whole population if we know only part of the data.
The term population refers to the whole collection of people or things
being considered. A sample is a subset of the total population, whether
people, auto parts, or anything else we are investigating. We want to
infer information about the whole population from information about the
sample.

Voluntary response polls can lead to results that present a totally
inaccurate view of the whole population. The simplest way to get a
sample that is likely to be representative of the whole population is to
just randomly pick a subset of the population. This is called a *simple
random sample* (SRS).

```{r}
coins <- \() sample(c("T", "H"), size=2, replace = TRUE)

experiment = purrr::map(1:1000, \(s) coins());
```

Ask everyone to raise their hand if they either threw a head or they
cheat.

```{r}
false_student <-  purrr::discard(experiment, \(s) s[1] == s[2])
true_student <-  purrr::keep(experiment, \(s) s[1] == s[2])
head_student <-  purrr::keep(experiment, \(s) s[1] == "H" &  s[2] == "H")

```

```{r}
counted <- list(false = length(false_student), true = length(true_student), head = length(head_student))
```

• Suppose we do this experiment with 1000 students, and 800 raise their
hands.

```{r}
cmStats$P(counted$head, counted$true)
```

We can estimate that about 60% of the students cheat because if 500
threw heads, then 300 of the 500 who threw tails cheated—about 60%
(300/500).

```{r}
cmStats$P(300, 500)
```

In taking samples, our goal is to obtain samples from which we can
accurately infer information about the whole population. That is, we
want our samples to be representative of the whole population. Pitfalls
to avoid in taking samples include avoiding biased samples and being
wary of results obtained from voluntary response samples. We can obtain
statistical information concerning sensitive information by using clever
techniques in which no 43 individual needs to divulge sensitive secrets.
The basic purpose of getting data from a sample is to infer information
about the whole population. How that inference is made is the subject of
the next lectures. \# Concept and Applications

## POINT ESTIMATES AND STANDARD ERROR

Descriptive statistics focus on summarizing characteristics about our
data, such as calculating the sample mean or plotting histograms, and
describe the dataset that’s being analyzed but don’t let us to draw any
conclusions or make any interferences about the data. On the other hand,
statistical inference uses our dataset to extract information about
populations or answer real-world questions with certainty. It builds on
the methods of descriptive statistics, but we can draw conclusions about
the population based on data from a sample.

### POINT ESTIMATES

Any time we have a random sample, we can calculate the sample mean,
variance, and standard deviation. These values are called point
estimates.

```{r}
head(Orange)
```

The dataset “Orange” has 35 rows and 3 columns of records of the growth
of orange trees. Find a point estimate for the average orange tree
growth and average tree circumference using the orange tree data.

```{r}
summary(Orange)
```

A point estimate for the average orange tree age would be 922.14 days,
and a point estimate for orange circumference would be 115.85
millimeters.

```{r}
as_tibble_row(list(age =  mean(Orange$age), circ = mean(Orange$circumference)))
```

The median is also a perfectly valid point estimate.

```{r}
as_tibble_row(list(age =  median(Orange$age), circ = median(Orange$circumference)))

```

Notice how close the sample mean circumference is to the sample median
of circumference: 115.85 compared to 115. And look how far off tree age
estimates are from each other: 922.1 compared to 1004.

Which estimate is better? How do we decide between the 2?

### STATISTICAL INFERENCE

Statistical inference is the process of making an estimate, prediction,
or decision about a population based on a sample.

The rationale is that large populations make investigating every member
impractical and expensive. It’s easier and cheaper to take a sample and
make estimates about the population from the sample. But those estimates
aren’t always going to be correct.

Statistical inference helps us do 2 main things: Draw conclusions about
a population or process from sample data and provide a statement of how
much confidence we can place in our conclusions.

### ESTIMATION

The objective of estimation is to approximate the value of a population
parameter on the basis of a sample statistic. We can estimate both
points and intervals.

A point estimate of a parameter θ is a single number that can be
regarded as a sensible value for θ. We find a point estimate by
selecting a suitable statistic and computing its value from given sample
data. The selected statistic is called the point estimator.

Point estimate is a value; point estimator is a statistic. Usually, we
use to denote the point estimator of a parameter θ.

The “women” dataset in R contains the average heights and weights for
American women aged 30 to 39. This is a sample of 15 women with those 2
variables: height in inches and weight in pounds.

```{r}
summary(women)
```

```{r}
head(women)
```

```{r}
plot(women)
```

```{r}
list(mean = mean(women$weight), median = median(women$weight))
```

We can generate point estimates for the population parameters based on
our sample statistics. What other estimators could we use to estimate
the population mean ( μ )?

Average of the extremes: . In fact, we’re taking the range and halving
it, so this gives us a measure of the center of the dataset.

Because we don’t know how close we are to the true population parameter,
we have to be strategic in selecting our point estimator. In other
words, we want our point estimate to follow certain rules so that we
know our sample estimate will be close to the population estimate.

### CHOOSING AN OPTIMAL ESTIMATOR

How do we choose an optimal estimator?

A point estimator is a random variable, and its value varies from sample
to sample. That variability is called the error of estimation.

```{r}
errro_of_estimation ~ phi:hat - phi
```

```{r}
# solving for phi_hat:
phi_hat ~ phi + error_of_estimation
```

Which is the best one? We want to choose an estimator that is unbiased,
one that is closest to the true value of the parameter θ on average, and
that has minimum variance, or the one with the smallest estimation of
error—the smallest variability, the least spread.

A point estimator is an unbiased estimator of the parameter θ if E( If
is not unbiased, the difference E( ) − θ is called the bias of.

Other estimates exist, such as the range, or the average of the largest
and smallest values. But these estimates are not unbiased and should not
be used to estimate the mean.

```{r}
cmStats$mu_sym
```

Sample variance $s^2$ the unbiased estimator of the $𝑛 − 1$, population
variance $σ2$.
```{r}
sum(cmStats$X - cmStats$mu_sym)^2/ (cmStats$n - 1)
```
```{r}
cmStats$σ_2_sym
```
Suppose that X is a binomial with parameters 𝑛 and 𝑝, with 𝑝 unknown.
We can show that the estimator = 𝑥 / 𝑛 is an unbiased estimator.

```{r}
E(X/n) ~ 1/n * E(x) ~1/n * np ~ p
```
For a normal distribution with mean μ and variance σ2, given a random
sample of size 𝑛, X1, X2, … X𝑛, we can show that the sample mean
is an unbiased estimator of μ.

```{r}
E(X_bar) ~ E(Sigma(X)/n) ~ 1/n * E(Sigma(X))
```

### CONSISTENCY

An unbiased estimator is consistent if the variance of the estimator
approaches 0 as our sample size ( 𝑛 ) approaches infinity. In other words,
the difference between the estimator and the population parameter
becomes smaller as the sample size increases.

As another example, suppose instead that we estimate the mean by
averaging the first and last values, X1 and X𝑛. This estimate is unbiased
because

```{r}
E((X_1 + X_n)/2) ~E(X_1)/2 + E(X_n)/2 ~ mu/2 + mu/2 ~ mu
```

```{r}
as_func(cmStats$μ_hat_sym)
#(min(1:10) + max(1:10))/2
sum(cmStats$x)
``` ### still not finis!!!!!

