---
title: "Statistics Everywhere"
output: html_notebook
---

# Foundational Concepts

Often, statistical reasoning can contribute to decisive arguments in
matters that seem very dif cult to resolve, and even issues that don’t
appear to have any statistical component to them at all.
```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)
library(quanteda)
library(DiscreteDatasets)
library(rstanarm);
```

```{r}
def_sym("X", "Y")

cmStats <- modules::use("CmStats.R")$cmStats

set.seed(3456)
```

Statistics is a subject that permeates essentially every area of our lives
and world. It is a powerful tool for seeing our world in a more detailed
fashion and for making informed decisions, although its subtleties and
potential misuses caution us to avoid thoughtless acceptance of statistical
conclusions. The recent and expected future development of computer speed
and capacity allow us to imagine using statistics with ever more scope and
effect. How much information and understanding can we hope to gather from
statistical data? How much more meaning would better statistical techniques
allow us to nd? Statistics is a tool with wide applicability. It has limits that
need to be acknowledged and respected, but its potential for helping us nd
meaning in our data-driven world is enormous and growing.

### Frequentist Stylometry Analysis

*This workflow uses the frequency of function words to distinguish between two authors (Hamilton and Madison), which is the core principle of the Federalist Papers analysis.*

This Federalist Papers example is satisfying in that it suggests that seeking
data to nd persuasive arguments is a valuable method for coming to
conclusions. Sometimes it is not clear what data are pertinent. In the case
of The Federalist Papers dispute, it would not be immediately obvious that
counting the frequency of trivial words would be the road to decision on the
authorship issue.

The evidence for the determination that
Madison wrote the anonymous articles
would be evaluated differently by Bayesian
statisticians versus frequentist statisticians. A
Bayesian would be willing to say that, given
the word usage in The Federalist Papers,
there is a 99.9% chance that it is written by
Madison. Frequentists would say that the
articles are either written by Madison or not. Both camps would probably
agree to a statement something like this: “The probability is only 2.4% of
getting no upons when randomly selecting a collection of 1000 words from a
person’s writing that generally has 6 upons per 1000.”

```{r}
ggplot(federalist, aes( x = doc_author, y = upon )) +
  geom_count()

```


```{r}

word_count_cols <- names(federalist)[3:ncol(federalist)]

federalist_processed <- federalist %>%
  mutate(
    # Sum the counts across all word columns for each paper
    total_words = rowSums(select(., all_of(word_count_cols)))
  ) |>
  select(doc_author, upon, total_words) |>
  rename(Author = doc_author) |>
  mutate(upon_rate = (upon / total_words) * 1000)

federalist_processed

```

```{r}
# 1. Filter for known authors (Hamilton and Madison)
known_papers <- federalist_processed |>
  filter(Author %in% c("Hamilton", "Madison"))
  # Rename for cleaner formula output
# 2. Perform the T-test
# H0: Mean usage rate of 'upon' is the same for both authors.
t_test_result <- t.test(upon_rate ~ Author, data = known_papers)

t_test_result
```
*The resulting p-value will provide the Frequentist conclusion on whether the word "upon" is a reliable fingerprint for attributing the disputed Federalist Papers.*

Using data and statistical analyses will become an even more prominent
part of our world in the future than it is now. The principal reason is the
continuing development of computer technology. With the computer, it is
now possible to deal with large databases and use techniques that would
have been computationally impossible previously. Some such techniques
involve simulation as a means to understand a collection of data. Often,
such methods are computationally intensive and, consequently, become
increasingly valuable as computer power increases.

### Monte Carlo method

One such technique is called the Monte Carlo method, which involves using
random processes by a computer to generate thousands of scenarios, enabling
statistical techniques to derive the distribution of the behavior of a system.
Today on a home computer, it is possible to do amazing statistical analyses
essentially instantly. Early textbooks on statistics would emphasize methods
for reducing computation. Now, with computers, those techniques are not
so important.

```{r}
known_papers_bayesian <- federalist_processed |>
  filter(Author %in% c("Hamilton", "Madison")) |>
  filter(!is.nan(upon_rate) & upon_rate > 0)
```

```{r}

bayesian_model_upon <- stan_glm(
  upon_rate ~ Author,
  data = known_papers_bayesian,
  family = gaussian(), 
  prior = normal(0, 5), # A weakly informative prior (standard for stylometry)
  chains = 2, iter = 1000
)

print(summary(bayesian_model_upon, digits = 3), pars = "beta")
```

```{r}
posterior_samples <- as_tibble(bayesian_model_upon)


mean_diff <- posterior_samples %>%
  summarise(
    MeanDifference = mean(AuthorMadison),
    ProbMadisonHigher = mean(AuthorMadison > 0)
  )

mean_diff
```
The output of the MCMC process (posterior_samples) is the simulated distribution of the parameter values.

- The MeanDifference tells you the posterior estimate of how much higher Madison's usage rate of "upon" is compared to Hamilton's.

- The ProbMadisonHigher (which will likely be close to 1.0) is the ultimate Bayesian conclusion: it's the probability that the rate of "upon" usage for Madison is truly higher than Hamilton's.

*The strength of the Bayesian Monte Carlo method is that it provides a direct probability statement $(P(Madison∣Data))$, rather than just a p-value $(P(Data∣H 0))$.*

### Schools of thought with common basics

We have seen statistics as having two basic parts:
- Organizing, describing, and summarizing a collection of data when we
know all the data.

- Inferring information about the whole population when we have data
about only a sample of the population.
When we make an estimate of the value of a feature of the whole population
given data about a sample, our challenge is to describe how accurate our
estimate is likely to be by answering the following questions:

- How close is our estimate to the correct value?

- How con dent are we that our estimate is, in fact, that close?



```{r}
tibble(Approach = c("Frequentist (Classical)", "Bayesian"), 
       Key_idea = c("The p-value. Probability is the long-run frequency. Parameters are fixed but unknown. Inference relies on data alone.", "Probability is degree of belief. Parameters are random variables. Inference updates prior beliefs with new data to form a posterior distribution."), 
       R_Focus = c("t.test(), lm(), chisq.test().", "rstanarm or brms packages"))
```

**Statistics is a powerful tool for understanding our world. We end the course
with a statistic about the course: Among people who learn something about
statistics, 100% appreciate our world with more clarity.**


# Concept and Applications

## STATISTICS YOUR WAY WITH CUSTOM FUNCTIONS

One of the coolest things you can do in R is write your
own functions. Custom functions allow you to define
a specific action you are interested in, which you
can then easily apply to new data. This can be anything
from calculating a unique statistic, to creating a custom
plot, to combining several outputs into a single display.

### WRITING CUSTOM FUNCTIONS

To make a custom function, you need to define one or more inputs
followed by some action and an output. Your function will take the input
and perform the action based on that input and spit out some output in
the form of transformed data, summary statistics, or a plot.

For example, let’s create a custom function that replicates the “mean( )” function in R:

```{r}
mean.fun <- function(x)
{
my.mean <- sum(x) / length(x)
return(my.mean)
}
```
Once you’ve defined a function and given it a name, you can then use the function on new data:
```{r}
mean.fun(0:20)
```
```{r}
mean.fun(rnorm(300))
```
While the function “mean.fun” only had 1 input ( the vector 𝑥 ), we can
also define multiple inputs. For example, let’s adjust the function so that
it has an input “delete.outliers,” which, when true, automatically deletes
outliers before calculating the mean. This will come in handy whenever
you do exploratory data analysis.

```{r}
mean.fun <- function(x, delete.outliers = F)
{
  if(delete.outliers == F) {
  my.mean <- sum(x) / length(x)
  }
  if(delete.outliers == T) {
  Orig.Mean <- sum(x) / length(x)
  Orig.SD <- sd(x)
  Outlier.log <- (x > Orig.Mean + 2 * Orig.SD) | (x <
  Orig.Mean - 2 * Orig.SD)
  x.noout <- x[Outlier.log == FALSE]
  my.mean <- sum(x.noout) / length(x.noout)
  }
  return(my.mean)
}
```

Here is an example of the function in action. Let’s create a vector “data”
with 100 samples from a standard normal distribution with 1 outlier
at 400.
```{r}
Data <- c(rnorm(n = 100, mean = 0, sd = 1),400)
mean.fun(Data, delete.outliers = F)
mean.fun(Data, delete.outliers = T)
```
Once we’ve defined a custom function in R, we can then use that
function in other functions.

### MODIFYING EXISTING FUNCTIONS

If you find that you’re frequently adding certain things to plots, such as
reference lines for the median or mean, then creating a custom plotting
function can make things easier for you.
```{r}
hist.fun <- function(x, add.median = T) {
hist(x)
if(add.median == T) {abline(v = median(x), lwd = 3, col
= "blue")}
}
```

Use “abline” to define any line, and then “v = ” for a vertical line.
```{r}
hist.fun(rexp(100), add.median = T)
```
Let’s modify our custom histogram function “hist.fun”:
```{r}
hist.fun <- function(x, add.median = T, add.mean = T,
add.legend = T, num.breaks = 12) {
b <- seq(min(x), max(x), length=num.breaks)
hist(x, col="cadetblue", breaks = b)
if(add.median == T) {abline(v = median(x), lwd = 3, col
= "blue")}
if(add.mean == T) {abline(v = mean(x), lty = 2, lwd = 3,
col = "red")}
if(add.legend == T) {
legend("topright", c("median", "mean"), lwd = c(3, 3),
lty = c(1, 2), col = c("blue","red"))
}
}
hist.fun(rexp(100), add.median = T,
add.mean = T, add.legend = T)
```
Let’s try this function on another dataset.

```{r}

hist.fun <- function(x, add.median = T, add.mean = T,
add.legend = T, num.breaks = 12) {
b <- seq(min(x), max(x), length=num.breaks)
hist(x, col="cadetblue", breaks = b)
if(add.median == T) {abline(v = median(x), lwd = 3, col
= "blue")}
if(add.mean == T) {abline(v = mean(x), lty = 2, lwd = 3,
col = "red")}
if(add.legend == T) {
legend("topright", c("median", "mean"), lwd = c(3, 3),
lty = c(1, 2), col = c("blue","red"))
}
}

hist.fun(mtcars$mpg, add.median = T, add.mean = T, add.legend = T)
```

We can generate plots with different breaks so that we can compare them.
```{r}
par(mfrow=c(1,3))
hist.fun(chickwts$weight,num.breaks = 5)
hist.fun(chickwts$weight,num.breaks = 20)
hist.fun(chickwts$weight,num.breaks = 30)

```

We can add as many custom procedures and plotting parameters as
we’d like. In fact, we could create a similar function for scatterplots.

We can create a custom scatterplot with some fun features.

```{r}
scatter.fun <- function(x = rnorm(1000), y = rnorm(1000),
add.mean = T) {
plot(x, y, pch=20)
if(add.mean == T) {
abline(h = mean(y))
abline(v = mean(x))
}
}
```

Let’s run it with defaults.
```{r}
scatter.fun()
```
Let’s see our custom scatterplot in action. We can create 2 plots, one where there is not significant correlation and one where there is.
```{r}
x <- rnorm(100)
y.uncorr <- rnorm(100)
y.corr <- x + rnorm(100, 0, .3)
par(mfrow = c(1, 2))
scatter.fun(x, y.uncorr)
scatter.fun(x, y.corr)
```

“Add.regression” adds a regression line. If the correlation is significant,
we'll make the line green; if not, it will be red.

```{r}
scatter.fun <- function(x = rnorm(1000), y = rnorm(1000),
  add.mean = T, add.regression = T,
  add.conclusion = T) {
  plot(x, y, pch=20)
  if(add.mean == T) {
  abline(h = mean(y))
  abline(v = mean(x))
  }
  if(add.conclusion == T) {
  C.Test <- cor.test(x, y)
  Coefficient <- C.Test$estimate
  p.value <- C.Test$p.value
  if(p.value <= .05) {
  Conclusion <- paste("The correlation coefficient is ",
  round(Coefficient, 2),
  ".\nIt IS significantly different from 0 (p =
  ", round(p.value, 3),
  ")", sep = "")
  }
  if(p.value > .05) {
  Conclusion <- paste("The correlation coefficient is ",
  round(Coefficient, 2),
  ".\nIt is NOT significantly different from 0
  (p = ", round(p.value, 3),
  ")", sep = "")
  }
  mtext(text = Conclusion,
  side = 3, line = 1, cex = .8) }
    
    if(add.regression == T) {
  Model <- lm(y ~ x)
  p.value <- anova(Model)$"Pr(>F)"[1]
  if(p.value <= 0.05) {Reg.Line.Col <- "green"}
  if(p.value > 0.05) {Reg.Line.Col <- "darkred"}
  abline(lm(y ~ x), lty = 2, lwd = 4, col = Reg.Line.Col)
  }
}
```

```{r}
par(mfrow = c(1, 2))
scatter.fun(x, y.uncorr)
scatter.fun(x, y.corr)
```

### PITFALL

- R has built-in functions, such as “summary( )” or “median( ),” and this
overlap is the most common pitfall: overwriting a built-in R function.

- Some of R’s built-in functions are common words, such as “contour,”
“image,” “jitter,” “length,” “mean,” “median,” “order,” “plot,” “sample,”
“sign,” “sequence,” “sort,” and “summary.”

**Before creating a new function, always type “?( new function name )” to
see if it already exists as an R function.**
