---
title: "Science—Mendel’s Too-Good Peas"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Foundational Concepts

In this lecture, we’re going to be talking about science and applications of statistics to science. Certainly, statistics and the statistical analysis of data are obviously central players in all aspects of science.

```{r}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
library(caracas)
library(sets)
library(tibble)
library(ggplot2)
library(dplyr)
library(data.table)
library(plotly)
library(RKaggle)
```

```{r}
def_sym("X", "Y", "a" , "b")

cmStats <- modules::use("CmStats.R")$cmStats

solar_system <-  RKaggle::get_dataset("cpitrat/solar-system-planets-motions");
```

Advances in empirical science depend on drawing deductions from data. In many cases, a scienti c theory is tested by comparing experimental results to predictions of the theory. Randomness can enter in two ways. First, measurement error (noise) in experimental results adds randomness to otherwise de nite predictions. Second, some theories, such as Mendel’s theory of trait inheritance over generations of pea plants, are inherently probabilistic. In fact, reported results with too little uctuation can be evidence of fraudulent data. On the other hand, a measurement much different from other measurements (an outlier) can indicate either that some gross error in measuring has occurred, in which case the measurement should be discounted, or that some fundamental assumption is incorrect. Study of the ozone layer in the atmosphere supplies a cautionary example.

Statistics is involved in essentially all scienti c matters, from weather reports to quantum physics. During the last 400 years, we humans have fundamentally altered our conception of the universe and our position in it, in many cases as a result of some scienti c advance that ultimately is based on the analysis of data. In this lecture, we’ll look at several examples of scientific developments and the role of statistics in them.

```{r}
head(solar_system)
```

The rst example involves Johannes Kepler, the famous astronomer, working in about 1600 as assistant to the astronomer Tycho Brahe, who had amassed vast amounts of data about the locations of planets and stars. Kepler computed that the data t a model of the solar system in which the planets revolve around the Sun following elliptical orbits. In devising his laws of planetary motion, Kepler used the statistical technique of creating a mathematical model to summarize the data.

```{r}
lazy_dt <-  dtplyr::lazy_dt(solar_system)

plot_data_efficient <- lazy_dt %>% 
  select(body, x, y, z) %>%
  mutate(body = factor(body)) %>%
  filter(body != "Sun") %>%
  as_tibble()
head(plot_data_efficient)
```

```{r}
# Plotting the 1D Orbits (X-Y Plane)
orbit_plot_2D <- ggplot(plot_data_efficient, aes(x = x, y = x, group = body, color = body)) +
  
  geom_path(linewidth = 0.5) +
  
  geom_point(aes(x = 0, y = 0), color = "orange", size = 5) +
    coord_fixed() +
  
  labs(
    title = "1D Projection of Planetary Orbits",
    x = "X Coordinate (AU or scaled units)",
    y = "Y Coordinate (AU or scaled units)"
  ) +
  theme_minimal()

orbit_plot_2D
```

Later, Isaac Newton formulated his universal law of gravitation, which implies that two masses will follow elliptical orbits about one another.

```{r}
orbit_plot_2d <-  ggplot(plot_data_efficient, aes(x = x, y = y, group = body, colour = body)) +
  geom_path(linewidth = 0.5) +
  geom_point(data = plot_data_efficient %>% group_by(body) %>% slice_tail(n = 1), size = 2) +
  geom_point(aes(x = 0, y = 0), color = "orange", size = 5) +
  geom_text(aes(x = 0, y = 0), label = "Sun", vjust = -1, color = "black") +
  
  coord_fixed() +
  
  labs(
    title = "2D Projection of Planetary Orbits (X-Y Plane)",
    x = "X Coordinate (AU or scaled units)",
    y = "Y Coordinate (AU or scaled units)",
    caption = "Data Source: Solar System Planets Motions Dataset"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
orbit_plot_2d
```

Science frequently progresses in this way. Observations are made that are well summarized by a mathematical equation or model, based on statistical curve-fitting techniques. Later, a more basic understanding of causes and effects can explain that physical model.

```{r}


# Use plot_ly and add_paths to trace the 3D orbits
plot_ly(plot_data_efficient, 
        x = ~x, 
        y = ~x, 
        z = ~x, 
        color = ~body, 
        type = 'scatter3d', 
        mode = 'lines'
) %>%
  layout(title = "3D Interactive Planetary Orbits",
         scene = list(aspectmode = "data")) # Keeps scales proportional
```

Another example in astronomy is the 3-degree radiation left over from the Big Bang. Researchers were trying to build a precise radio telescope and kept sensing background noise. After many attempts to ne-tune their instruments to avoid that “error,” the researchers discovered that the background noise was a real phenomenon: the 3-degree Kelvin radiation left over from the Big Bang at the creation of the universe.

Randomness is at the heart of quantum physics. Modern theories of physics postulate the very unintuitive concept that a subatomic particle, such as an electron, is not in a precise location at a particular time. Instead, the location of an electron is a probability distribution. These theories put the statistical and probabilistic nature of existence in a fundamental position in our understanding of the world.

Another example of statistics in science is Mendel’s famous experiments with peas. Mendel noted statistical patterns in data concerning hereditary traits of pea plants. Yellow is the dominant gene. If homozygous yellow pea plants (those with two yellow genes) and homozygous green pea plants (those with two green genes) are crossed, the rst-generation offspring all look yellow, being yellow heterozygous plants (those with one yellow gene and one green gene). The second-generation offspring, however, which are the result of crossing yellow heterozygous plants, are about one-quarter green, indicating homozygous green pea plants.

```{r}
observed_counts <- c(Round = 428, Wrinkled = 152)
total_count <- sum(observed_counts)


expected_proportions <- c(Round = 3/4, Wrinkled = 1/4)

expected_counts <- total_count * expected_proportions
expected_counts
```

Statisticians have looked at Mendel’s reported results and have discovered that it would be very unusual, given the amount of his data, that all of the results would be in the narrow bounds he reported. In short, Fisher believed that the data that Mendel got were too good. Ronald Fisher, to whom we were introduced in Lecture 12, found that Mendel’s results lie within 1 standard deviation of the mean much more often than the expected 68% of the time.

```{r}
mendel_chi_sq_test <- chisq.test(
  x = observed_counts, 
  p = expected_proportions
)

mendel_chi_sq_test
```

Fisher also noted that Mendel used the method of cross-breeding yellow plants with themselves 10 times to identify whether they were homozygous or heterozygous. In an interesting twist, this method implies that we should have expected Mendel to misclassify a certain percentage of plants; however, Mendel’s reported data are closer to the data expected if all the plants were classi ed correctly. Mendel’s work illustrates all aspects of statistics, including design of experiments and interpretation of data, and may illustrate the possibility that the data were made to look somewhat better than they actually were.

Using carefully executed statistical capture-recapture methods, scientists can estimate quantities as diverse as the population of tigers in a jungle, the volume of water in a lake, and the size of a natural gas deposit in the ground. Statistical analysis of experimental data is key to validating or invalidating a scienti c theory.

# Concept and Applications

## PRIOR INFORMATION AND BAYESIAN INFERENCE

Thus far, our approach to statistical inference has been greedy for new data. Getting more new data, and increasing our sample size, are the keys to making our inferences even more reliable. The approach we have been following is called a frequentist approach. But what about also using prior data and prior information? This is the central idea of Bayesian inferential statistics, which gives us a way to update our prior beliefs based on observed data.

### BAYESIAN STATISTICS

Bayesian statistics is an entirely different approach for doing statistical inference. In other words, it’s not just another technique like regression, hypothesis testing, or ANOVA.

-   The methods are based on the idea that before we ever observe data, we have some prior belief about it, perhaps based on experience or other experiments. As we observe new data, we use the data to update our belief.

-   A lot of what you’ve learned in this course stays the same with Bayesian statistics. Histograms, box plots, and numerical summaries stay largely the same. Discrete and continuous distributions, along with special cases, such as the binomial, exponential, uniform, and normal distributions, stay the same, but we often make more use of a variety of distributions. Uncovering a relationship between 2 variables in terms of correlation and covariance also stays much the same.

-   But the way that we have been drawing conclusions about a population of interest through t-tests and hypothesis testing looks very different in a Bayesian approach to inference.

-   In a frequentist approach, we worked hard to obtain a random sample, modeled the linear relationship between response and explanatory variables, and attached measures of confidence to them.

**The Bayesian approach is useful for estimating a proportion, where you have a rough idea but you don’t have much data. Your uncertainty is described by a probability distribution called your prior distribution.**

*In Bayesian statistics, the terms in Bayes’s rule have special names.*

.1 The unknown parameter θ is considered a random variable—no longer fixed. We use a probability distribution ( normal, uniform, etc. ) to describe the uncertainty of the unknown parameter.

.2 P( θ ) is the prior probability, which describes how sure we were that θ was true before we observed the data.

.3 Once we have our data, we update the parameter distribution using Bayes’s theorem.

.4 P( θ\|data ) is the posterior probability. It describes how certain or confident we are that the hypothesis is true, given that we have observed data.

Calculating posterior probabilities is the main goal of Bayesian statistics. P( data\|θ ) is the likelihood. If you were to assume that θ is true, this is the probability that you would have observed the data. 𝑝( data ) is the marginal likelihood. This is the probability that you would have observed data whether θ is true or not. Using Bayes’s rule:

```{r}
P(Ø | X) ~ P(X | Ø)*P(Ø)/P(X)

```

```{r}
posterior_of(Ø) ~ Inf ~ prior_of(Ø) * likehood
```

.5 The Bayes estimator of θ is either the mean or the median of the posterior distribution—you get to choose. In Bayesian statistics, both the mean and median are valid estimators.

-   The absolute loss function is as follows: $L( θ, ) = |θ − |$.

```{r}
cmStats$abs_loss_sym
```

-   The squared loss function is as follows: $L( θ, ) = ( θ − )^2$

```{r}
cmStats$sqrt_loss_sym
```

. - For squared error loss, the posterior mean is the Bayes estimator.

-   For absolute error loss, the posterior median is the Bayes estimator.

### BETA BINOMIAL DISTRIBUTION

-   The beta distribution is a common distribution that is used to describe our prior knowledge. It can be understood as representing a distribution of probabilities—that is, it represents all the possible values of a probability when we don’t know what that probability is.

-   In baseball, a batting average is the number of times a player gets a base hit divided by the number of times the player goes up at bat. The batting average lies between 0 and 1; a decent batting average ranges between 0.25 and 0.27 while a batting average over 0.300 is excellent.

-   If we’re putting together a fantasy baseball league, we might want to predict the season’s batting average for a player at the beginning of the season. We have some prior information about batting averages: Most season averages range between 0.220 and 0.365.

-   There’s a binomial distribution lurking in the background, because at its core, the batting average represents a series of successes ( hits ) and failures ( misses ).

The domain of the beta distribution is ( 0, 1 ), which lines up with the appropriate range for our batting average. We expect that the player’s season-long batting average will be most likely around 0.26 but that it could reasonably range from 0.20 to 0.36. This can be represented with a beta distribution with parameters 𝑎 = 78 and 𝑏 = 222:

```{r}
curve(dbeta(x, 78, 222))
```

Assuming a season with 300 at bats, 𝑎 = 78 represents the number of hits and 𝑏 = 222 represents the number of misses. The mean of the beta distribution is $𝑎/( 𝑎 + 𝑏 ) = 78/( 78+222 )$. And the distribution pretty accurately contains our prior belief of a batting average between ( 0.20, 0.36 ).

```{r}
78/( 78+222 )
```

The new beta distribution will be beta( 𝑎0+ hits, 𝑏0 + misses ), where 𝑎0 and 𝑏0 are the parameters we started with—that is, 78 and 222.

In this case, 𝑎 increases by 1 ( the player’s 1 hit ) while 𝑏 has not increased at all ( no misses yet ). That means that our new distribution is beta( 78 + 1222 ), or:

```{r}
curve(dbeta(x, 79, 222))
```

You can’t really tell the difference here, but there was a small shift. Suppose that halfway through the season the player has been up to bat 300 times, hitting 105 out of those times. The new distribution would be beta( 79 + 105, 222 + 195 ).

```{r}
curve(dbeta(x, 79+105, 222+195))
```

The curve is now thinner and slightly shifted to the right to reflect the higher batting average. In fact, the new expected value is our posterior estimate of the player’s batting average. We can calculate it as $𝑎/( 𝑎 + 𝑏 )$.

After 105 hits of 300 real at bats, the expected value of the new beta distribution is $𝑎/( 𝑎 + 𝑏 ) = ( 79 + 105 )/( 79 + 105 + 222 + 195 )$.

```{r}
( 79 + 105 )/( 79 + 105 + 222 + 195 )
```

Thus, the beta distribution is best for representing a probabilistic distribution of probabilities—the case where we don’t know what a probability is in advance but we have some reasonable guesses.

The Bayesian framework allows us to dynamically update our prior beliefs by considering various types of distributions for our prior probability.

In the absence of other information, a uniform ( or constant ) prior is often assumed, meaning that every possible value has the same probability of being selected.

-   Recall that a random variable has the uniform ( 0, l ) distribution if its probability density function is constant over the interval [0, 1] and 0 everywhere else.

```{r}
tibble(du = dunif(0:1, 0, 100)) |> ggplot(aes(x = 1:0, y = du, )) +
  geom_line() +
  ggtitle("Uniform Prior") +
  xlab("index")
```

```{r}
# Use the segments() function to draw a single line from (0, 1) to (1, 1)

# 1. Start a minimal plot canvas with the correct viewing limits
plot(x = 0, y = 0, type = "n", 
     xlim = c(0, 1), ylim = c(0.9, 1.1),
     axes = FALSE, xlab = "", ylab = "", main = "") # Hide everything but the data space

# 2. Draw the single constant line segment
segments(x0 = 0, y0 = 1, 
         x1 = 1, y1 = 1, 
         lwd = 3, col = "black")
```

On the other hand, the beta ( 𝑎, 𝑏 ) distribution is another commonly used distribution for a continuous random variable that can only take on values [0, 1].

It has the probability density function

```{r}
(factorial(a + b)/(factorial(a) * factorial(b))) * (x^(a - 1)) * (1 - cmStats$x)^(b - 1) 
```

where 𝛤( 𝑛 ) = 𝑛 − 1! and 0 \< 𝑥 \< 1.

```{r}
x ~ reals(l=0, r=1)

cmStats$factorial_sym(cmStats$n - 1)
```

```{r}
```

-   In the case of a binomial likelihood, any beta prior that we choose will give us a beta posterior distribution. When a prior and posterior belong to the same distribution family, this distribution is referred to as a conjugate prior. In this case, the beta distribution is a conjugate prior for the binomial likelihood.

-   Conjugate priors are immensely useful because they provide a simple analytic solution to this type of inference problem. We can easily know the mean and variance, for example. ۧ ۧ But conjugate priors are also somewhat limiting because our prior belief may not be representable using the conjugate family’s parameterization.

### PITFALLS

-   One of the pitfalls of Bayesian statistics is that 3 different people can choose 3 different priors, analyze the same dataset, and get 3 different posteriors.

-   A good ( informative ) prior will typically be based on a very similar study done under similar conditions in the past. But these aren’t always available, or applicable, in every situation. It’s also not as easy to write out a posterior distribution that doesn’t come from a conjugate family.

-   Suppose that your data followed a beta distribution and you wanted to use an exponential prior. The posterior would be a beta times an exponential distribution. We don’t know the mean or variance of that distribution. It’s actually more computationally intensive to try to estimate.

-   On the other hand, a Bayesian would complain that frequentist assumptions about long-run testing over an infinite number of hypothetical repetitions of the experiment may sound attractive but are not realistic. Bayesian methods better approach how we already think and the types of problems we often encounter.
